Causal Role of the Frontal Eye Field in Attention-induced Ocular Dominance Plasticity

Version 1

Abstract

Abstract

Previous research has found that prolonged eye-based attention can bias ocular dominance. If one eye long-termly views a regular movie meanwhile the opposite eye views a backward movie of the same episode, perceptual ocular dominance will shift towards the eye previously viewing the backward movie. Yet it remains unclear whether the role of eye-based attention in this phenomenon is causal or not. To address this issue, the present study relied on both the functional magnetic resonance imaging (fMRI) and transcranial magnetic stimulation (TMS) techniques. We found robust activation of the frontal eye field (FEF) and intraparietal sulcus (IPS) when participants were watching the dichoptic movie while focusing their attention on the regular movie. Interestingly, we found a robust effect of attention-induced ocular dominance shift when the cortical function of vertex or IPS was transiently inhibited by continuous theta burst stimulation (cTBS), yet the effect was significantly attenuated to a negligible extent when cTBS was delivered to FEF. A control experiment verified that the attenuation of ocular dominance shift after inhibitory stimulation of FEF was not due to any impact of the cTBS on the binocular rivalry measurement of ocular dominance. These findings suggest that the fronto-parietal attentional network is involved in controlling eye-based attention in the “dichoptic-backward-movie” adaptation paradigm, and in this network, FEF plays a crucial causal role in generating the attention-induced ocular dominance shift.

Introduction

Selective attention allows for the selection of pertinent information from a vast array of input for further cognitive processing (Treisman, 1969; Wolfe et al., 1989). It has been well known that attentional selection can be location-specific (Posner, 1980), feature-specific (Corbetta et al., 1990), or object-specific (Egly et al., 1994). Notably, selective attention can also be based on the eye of origin for visual input, named eye-based attention (Neisser & Becklen, 1975). For instance, presenting top-down attentional cues to one eye can intensify the competition strength of input signals in the attended eye during binocular rivalry (Choe & Kim, 2022; Zhang et al., 2012) and shift the eye balance towards the attended eye (Wong et al., 2021).

It is worth noting that eye-based attention not only exerts real-time effects on information processing (Choe & Kim, 2022; Wong et al., 2021; Zhang et al., 2012), but also exhibits a counterintuitive perceptual aftereffect that prolonged attention to a monocular pathway (or eye-based attention) can result in a shift of ocular dominance towards the unattended eye (Song, Lyu, Zhao, et al., 2023; Wang et al., 2021). In Song, Lyu, Zhao, et al. (2023)’s “dichoptic-backward-movie” adaptation paradigm (see Figure 1B), participants are presented with regular movie images in one eye (i.e. attended eye) while the other eye (i.e. unattended eye) received the backward movie images of the same episode. Song, Lyu, Zhao, et al. (2023) found that the predominance of the unattended eye in binocular rivalry increased after one hour of adaptation to the “dichoptic-backward-movie”, indicating a shift of perceptual ocular dominance towards the unattended eye. Furthermore, the steady-state visually evoked potentials (SSVEPs) elicited by the adapting stimuli in each eye were observed in both the occipital and frontal sites (Song, Lyu, Zhao, et al., 2023), with a significant positive correlation between the frontal neural-activity index (a normalized SSVEP response related to selective attention) and the shift of perceptual ocular dominance. Their findings imply the contribution of frontal attentional system in producing such an adaptation effect.

Schematic diagram of the (A) binocular rivalry task, (B) “dichoptic-backward-movie” paradigm and (C) blob target stimulus (see the gray region around the mouth in this example). This figure presents identifiable images of human faces solely for the purpose of demonstration, which were captured from the authors (F. Song, J. Wang, and J. Zhao) of this article. The movie images used in the experiment are not displayed in this figure due to potential copyright issues.

However, this account is based on the correlation result rather than causal evidence. Besides, the frontal activation observed by Song, Lyu, Zhao, et al. (2023) is primarily located in the frontal pole, which is not considered a key area directly responsible for attentional control (Iidaka, 2017; Jääskeläinen et al., 2016; Leminen et al., 2020; Xie et al., 2018). As SSVEP is limited to detecting brain activities that oscillate at the flickering frequency (and its harmonics) of the visual stimulus, brain activities that do not precisely align with the flickering frequency may be missed (Norcia et al., 2015). Consequently, the activities of typical brain regions associated with the attentional network, such as the parietal and dorsolateral prefrontal cortex (Vossel et al., 2014), may not be captured by the SSVEP measures.

Therefore, the present study combines the use of fMRI and TMS to further investigate the neural mechanisms underlying the perceptual aftereffects led by “dichoptic-backward-movie” adaptation, seeking direct causal evidence for the role of eye-based attention in modulating ocular dominance. In Experiment 1, we used fMRI to identify any brain regions associated with eye-based attention when subjects were watching the dichoptic movie. Subsequently, in Experiment 2, we examined whether transiently suppressing the cortical function of the identified brain regions using cTBS (Huang et al., 2005) would impair the formation of ocular dominance shift induced by “dichoptic-backward-movie” adaptation. Participants watched the dichoptic movie for 50 minutes after the delivery of cTBS, with their ocular dominance measured before and after adaptation by a binocular rivalry task. The fundamental logic of Experiment 2 can be summarized as follows. Firstly, the brain regions identified in Experiment 1 are related to eye-based attentional control, but do not necessarily contribute to the formation of ocular dominance shift. Secondly, if any of the identified brain regions plays a causal role in the attention-induced ocular dominance plasticity, transiently suppressing its cortical function will sharply weaken the magnitude of ocular dominance shift as compared to delivering the cTBS over vertex, a common control site in TMS studies. Thirdly, for a brain region with a causal role in generating the ocular dominance shift, whether suppression of its cortical function can completely eliminate the effect of ocular dominance shift or not may depend on the depth or efficiency of the suppression led by cTBS. In other words, even if the effect of ocular dominance shift does not vanish after the brain region is suppressed, as long as the effect becomes weakened, the brain region is still thought to play a causal role in the attention-induced ocular dominance plasticity. Experiment 3 further confirmed that the delivery of cTBS affected the formation of attention induced ocular dominance plasticity during adaptation rather than the performance of binocular rivalry task.

Results

Experiment 1: The function role of fronto-parietal areas in controlling eye-based attention

We first tried to identify any brain regions associated with eye-based attention using fMRI in Experiment 1. During the experimental runs of fMRI scanning, participants viewed movie images either dichoptically (Figure 2A) or binocularly (Figure 2B) through the paper red-blue stereoscopic glasses. In the dichoptic condition, regular movie images were presented to one eye while the other eye received the backward movie images of the same episode. In the binocular condition, the same regular movie images were presented to both eyes. Participants were instructed to attentively watch the regular movie and comprehend the plot. Besides, we compared the response difference between dichoptic and binocular conditions in the experimental runs with that in the control runs, where participants performed a central rapid serial visual presentation (RSVP) task while watching the movie stimuli (Figure 2C), to eliminate potential interference arising from different visual inputs between the dichoptic and binocular conditions.

The schematic of stimuli in (A) dichoptic condition or (B) the binocular condition in the experimental runs. (C) The schematic of RSVP task in the control runs. For demonstration purpose, the letters and fixation points are enlarged. (D) Schematic illustration of the generation process for red-blue movie images employed in the dichoptic condition.

Behavior result

After each experimental run, participants were required to answer a question about the movie plot. The results showed a high response accuracy to the questions (M = 91.67%, SD = 14.91%), indicating that the participants had paid close attention to the regular movie and clearly understood the plot.

In the control runs, participants also exhibited a reasonable performance in the RSVP task (d’ = 3.33, SD = 0.83). A 2 (experiment condition: dichoptic, binocular) × 3 (run: run1, run2, run3) repeated measurements analysis of variance (ANOVA) was conducted on the d-prime score, which indicated no significant main effect or interaction (ps > .09). The generally favorable RSVP performance across both experimental conditions suggested that participants faithfully focused their attention on the central letter stream without being much affected by the movie stimuli in the control runs. Otherwise, the RSVP performance would be found better in the binocular condition than in the dichoptic condition, because a disobedient participant who liked to complete the RSVP task and watch movie simultaneously had to pay more attention to the regular movie images (thus less attention to the letter stream) in the dichoptic condition than in the binocular condition given the annoying bistable nature of dichoptic stimulus presentation.

fMRI Result

Since only in the experimental runs that participants had to selectively pay more attention to the eye viewing the regular movie, we speculate that if certain brain regions exhibit greater activities in the dichoptic condition as compared to the binocular condition in the experimental runs but not in the control runs, the activation of these brain regions could be attributable to eye-based attention. After comparing the ‘dichoptic-binocular’ contrast between the experimental and control runs, we found 4 clusters that showed stronger responses to the dichoptic movies than to the binocular movies especially in the experimental runs (Figure 3). They located in the right FEF, bilateral IPS, and the right superior parietal lobule (SPL). In the control runs, the responses of these areas to the dichoptic and binocular movies showed no discernible difference. As previous research has demonstrated that these brain areas are part of the dorsal attentional network (Vossel et al., 2014), we therefore speculate that the identified clusters would be responsible for eye-based attention.

Illustration of the clusters with stronger ‘dichoptic-binocular’ contrast in the experimental runs than in the control runs. They were located in the right frontal eye field (FEF), bilateral intraparietal sulcus (IPS), and the right superior parietal lobule (SPL). The green circles indicate the corresponding cluster. The MNI coordinates represent the locations of peak voxels of each cluster. The color bar denotes the difference of ‘dichoptic-binocular’ contrast (β values) between the experimental and control runs.

Experiment 2: Suppressing FEF with cTBS attenuated the ocular dominance shift induced by eye-based attention

The results of Experiment 1 suggest that areas within the fronto-parietal network are responsible for eye-based attention. To further ascertain the causal role of eye-based attention in reshaping the ocular dominance, we then measured the effect of “dichoptic-backward-movie” adaptation (Song, Lyu, Zhao, et al., 2023) on ocular dominance with the function of any identified fronto-parietal area inhibited by cTBS. If suppressing any of the fronto-parietal regions results in a reduction or disappearance of ocular dominance shift after adaptation, we can infer that eye-based attention plays a causal role in the so-called attention-induced ocular dominance plasticity (Song, Lyu, Zhao, et al., 2023).

Participants completed a binocular rivalry task for the measurement of their ocular dominance before (pre-test) and after adaptation (post-test) (Figure 4A). cTBS was delivered over one of the areas identified in Experiment 1 (i.e. right FEF, right IPS) or a control region (vertex, (Wang et al., 2020)) after the pre-test (Figure 4B). Then, participants adapted to the “dichoptic-backward-movie” and were asked to detect some infrequent blob targets presented on the movie images in one eye at the same time.

Illustration of (A) the process in Experiment 2 and (B) the stimulation sites of cTBS.

Binocular rivalry test

To quantify perceptual eye dominance, we calculated the ocular dominance index (ODI) based on binocular rivalry dynamics. An increase in the value of ODI signifies a shift of ocular dominance towards the unattended eye. A 2 (test phase: pre-test, post-test) × 3 (stimulation site: Vertex, FEF, IPS) repeated measures ANOVA on the ODI (Figure 5A) showed a significant main effect of test phase (F (1,15) = 5.12, p = .039, η2 = .26) but not the main effect of stimulation site (F (2,30) = 0.24, p = .79, η2 = .02)), suggesting the ODI in the post-test (M = 0.49, SE = 0.02) was greater than that in the pre-test (M = 0.46, SE = 0.01, 95% CI = [0.002 0.053]).

The results of (A) the ocular dominance index (ODI), (B) the ocular dominance shift effects (ODSE) in Experiment 2, and (C) the ODI in Experiment 3. The bars show the grand average data for each condition. The individual data are plotted with gray lines or dots. The dashed gray line represents the absolute balance point for the two eyes (ODI = 0.5). Error bars indicate standard errors of means. * p < .05; ** p < .01; n.s. p > .05.

More importantly, the interaction between the test phase and the stimulation site was significant (F (2,30) = 3.67, p = .038, η2 = .20). Post-hoc test revealed that after adaptation a significant bias of ocular dominance in favor of the unattended eye could be observed if cTBS was delivered to vertex (ODIpre = 0.451±0.011; ODIpost = 0.494 ±0.015; t (15) = 3.76, p = .002, d = 0.94, 95% CI = [0.019 0.067], pFDR = .006) and IPS (ODIpre = 0.461±0.008; ODIpost = 0.486±0.014; t (15) = 2.39, p = .030, d = 0.60, 95% CI = [0.003 0.049], pFDR = .045), but not FEF (ODIpre = 0.463±0.008, ODIpost = 0.477±0.019; t (15) = 0.74, p = .47, d = 0.19, 95% CI = [-0.025 0.051]), indicating a sharp reduction of aftereffect for stimulation at the FEF only.

We further compared the magnitude of ocular dominance shift across different stimulation sites. For this goal, we computed the ocular dominance shift effect (ODSE) by subtracting the ODI in the pre-test from that in the post-test, then compared the ODSE between the stimulation sites using paired sample t-tests. As shown in Figure 5B, the ODSE after stimulating the vertex (M = 0.043, SE = 0.011) was greater than that after stimulating the FEF (M = 0.013, SE = 0.018; t (15) = 2.76, p = .015, d = 0.69, 95% CI = [0.007 0.053], pFDR = .044). No significant difference was found between other pairs (vertex vs. IPS, M = 0.026, SE = 0.011; t (15) = 1.70, p = .11, d = 0.43, 95% CI = [-0.004 0.039], pFDR = .17; FEF vs. IPS, t (15) = −1.05, p = .31, d = 0.26, 95% CI = [-0.038 0.013]).

Blob detection test

For each experimental condition, the target detection rate was calculated by dividing the summed number of detected blob targets by the total number of blob targets. A 2 (eye: attended eye, unattended eye) × 3 (stimulation site: Vertex, FEF, IPS) repeated measures ANOVA on the detection performance showed a significant main effect of eye (F (1,15) = 112.65, p < .001, η2 = .88) but no significant main effect of stimulation site (F (2,30) = 1.79, p = .18, η2 = .11) or the interaction (F (2,30) = 0.26, p = .78, η2 = .02), suggesting that the overall detection performance for the attended eye (M = 0.89, SE = 0.04) was better than that for the unattended eye (M = 0.23, SE = 0.06, 95% CI = [0.53 0.79]). These results confirm that more attention was allocated to the monocular pathway for the attended eye, with the interocular difference of detection rate unaffected by the cTBS.

Experiment 3: cTBS affected ocular dominance plasticity rather than the task performance reflecting ocular dominance

The results of Experiment 2 support the notion that eye-based attention was the cause for attention-induced ocular dominance plasticity. However, an alternative account is that the significant two-way interaction between test phase and stimulation site did not stem from any persistent malfunction of FEF in modulating ocular dominance, but rather it was due to some abnormality of binocular rivalry measures in the post-test that occurred after stimulation at the FEF only (and not at the other two brain sites). For instance, stimulation at the FEF might simply reduce the ODI measured in the binocular rivalry post-test.

An inspection of Figure 5A reveals that the ODI increased after the adaptation phase preceded by cTBS at the vertex, with most individual data points remaining below 0.5 (the absolute balance point for the two eyes) in the post-test. In other words, even after adaptation, most participants’ unattended eyes were still slightly less predominant than their attended eyes in binocular rivalry. Considering that vertex is a common control site in TMS studies, it can be inferred that a result pattern similar to that of the vertex would be observed if the FEF was not inhibited (i.e. ODI would also increase after adaptation yet still be smaller than 0.5). Hence, if stimulating the FEF simply influenced the binocular rivalry post-test by reducing the post-adaptation ODI, the ODI would then be further lowered along the y axis in Figure 5A, that is, the ocular dominance would become even more imbalanced after the stimulation. However, to our knowledge, TMS has been frequently found to rebalance both eyes rather than do the contrary (Clavagnier et al., 2013; Tuna et al., 2020; Tuna et al., 2022). Therefore, instead of abnormal measurement in the binocular rivalry post-test, we believe, the results of Experiment 2 more likely reflect a failure of ocular dominance shift towards the unattended eye due to a malfunction of FEF in adjusting ocular dominance during adaptation.

Despite all this, we still conducted Experiment 3 to examine how suppression of the three target sites would impact binocular rivalry performance, in case that any unknown confounding factors, unrelated to adaptation but related to binocular rivalry measures, contributed to the results. The task procedure was similar to that in Experiment 2 except that participants watched identical regular movie images with both eyes and were no longer required to perform the blob detection task. Since there was no dichoptic-backward-movie adaptation in this experiment, we expected no shift of ocular dominance at least for stimulating the vertex.

A 2 (test phase: pre-test, post-test) × 3 (stimulation site: Vertex, FEF, IPS) repeated measures ANOVA on the ODI showed neither a significant main effect nor the interaction (Figure 5C, ps > 0.43). Furthermore, in order to further examine these null effects, we conducted the Bayesian repeated measures ANOVA using JASP with default priors and computed inclusion Bayes factors (BFincl) which suggest the evidence for the inclusion of a particular effect calculated across matched models. Specifically, the Bayesian ANOVA provided moderate evidence (BFincl = 0.28, that is to say a BF between 1/10 and 1/3) to support the null hypothesis of no main effect of stimulation site and weak evidence (BFincl = 0.39, that is to say a BF between 1/3 and 1) in support of the null hypothesis of no main effect of test phase. Additionally, the Bayesian ANOVA yielded moderate evidence (BFincl = 0.23) supporting the null hypothesis of no interaction effect between the test phase and the stimulation site. These results suggested that the suppression of the fronto-parietal cortex had no impact on performance in binocular rivalry task.

Discussion

The present study investigated the neural mechanisms underlying the shift of ocular dominance induced by “dichoptic-backward-movie” adaptation. By using fMRI-guided TMS, we provide direct causal evidence for the effect of eye-based attention in modulating ocular dominance, and disclose the crucial role of FEF.

To our knowledge, our Experiment 1 for the first time explored the neural mechanisms of top-down eye-based attention. As compared to when participants watched a movie binocularly, we found stronger activation in FEF, IPS and SPL when they watched the same movie presented in one eye while ignoring the backward movie in the other eye. Since focusing on the movie in one eye necessitates more attention allocated to stimuli in that eye, the activation of these fronto-parietal areas likely reflects their functions in controlling eye-based attention. Given that the activated areas belong to the dorsal attentional network (Vossel et al., 2014), we advocate that top-down eye-based attention is also controlled by the dorsal attentional network. Furthermore, the activation for eye-based attention was predominantly localized in the right hemisphere, showing a typical right-hemisphere functional dominance of attentional control (Duecker et al., 2013; Mayrhofer et al., 2019; Sack, 2010).

Studies on the real-time effect of eye-based attention has shown that top-down eye-based attention can promote the attended eye’s dominance (Wong et al., 2021; Zhang et al., 2012). Yet both recent (Song, Lyu, Zhao, et al., 2023; Wang et al., 2021) and the present work has suggested that prolonged eye-based attention can lead to a counterintuitive perceptual aftereffect, a shift of ocular dominance towards the unattended eye. Interestingly, in our Experiment 2 this aftereffect was significantly attenuated after we temporarily inhibited the cortical function of FEF via cTBS. This finding indicates the crucial role of FEF in the formation of attention-induced ocular dominance shift.

To exclude the possibility that the cTBS on FEF simply affected the binocular rivalry measures, we conducted Experiment 3, the procedure of which resembled that of Experiment 2 except that participants were presented with regular movie images binocularly. This modification eliminated any involvement of eye-based attentional allocation. The results indicated that the cTBS delivered to FEF did not produce any significant changes on the binocular rivalry performance, thus supporting the notion that inhibition of the FEF undermined its modulatory role in ocular dominance plasticity during adaptation, which, in turn, caused a substantial reduction of ocular dominance shift towards the unattended eye.

Although both the FEF and IPS were found to be responsible for top-down eye-based attention in Experiment 1, the delivery of cTBS to IPS did not affect the adaptation-induced ocular dominance shift. Indeed, both areas belong to the dorsal attentional network (Vossel et al., 2014), yet previous researches have shown some functional distinctions between the two areas. For instance, it is found that the FEF plays a pivotal role in filtering out distractors (Lega et al., 2019) and sustaining attention (Esterman et al., 2015), whereas no such effect has been observed in the IPS. Furthermore, stimulation of the FEF (Ruff et al., 2006; Veniero et al., 2021) and IPS (Ruff et al., 2008) can both modulate activities in early visual areas, but produce qualitatively different effects. In this vein, our work reveals a new type of functional distinction between the two areas, showing a unique role of the FEF in attention-induced ocular dominance plasticity.

One may be confused why cTBS did not impair the performance of blob detection for the attended eye, given that the interocular difference of blob detection rate may be a coarse estimation of the real-time effect of eye-based attention. Let us recall our findings in the previous work (Song, Lyu, Zhao, et al., 2023): 1) The magnitude of aftereffect (i.e. attention-induced ocular dominance shift) correlated with the frontal but not occipital neural-activity index that was calculated as the ratio of neural response for the attended eye to that for the unattended eye; 2) the high-demanding “No-Sound” condition yielded even smaller interocular difference of detection rate than the “Synchrony” condition (where the audio track was presented and synchronized with the regular movie, like Experiment 2 of the present work) but created the greatest magnitude of aftereffect among all adapting conditions; 3) the aftereffect could be produced only when the adapting stimuli involved adequate interocular competition, and also measurable only when the testing stimuli were not binocularly fused. These results suggest that the aftereffect may not be tightly associated with the real-time effect of eye-based attention, but mainly dependent upon the degree of involvement of attentional control mechanisms on the premise of interocular competition. We then speculated that the aftereffect might reflect a modulation of the neurons that gated interocular competition (Katyal et al., 2018; Said & Heeger, 2013; Song, Lyu, & Bao, 2023) by the frontal attentional system (Song, Lyu, Zhao, et al., 2023). From this perspective, inhibition of the FEF in our Experiment 2 might undermine the efficacy of such attentional modulation to an extent that sufficed to prevent a change of ocular dominance. Yet with the help of synchronized audio, residual attentional control following the cTBS might still be competent to sustain attention to the regular movie, allowing a sound performance for this relatively easy task (detection targets appeared only once every 5 min in each eye). Alternatively, inhibition of the FEF could destroy its functional feedbacks to the neurons that gated interocular competition, but its function in controlling eye-based attention is either spared or compensated by IPS or other regions in the dorsal attentional network, which avoids a significant downgrade of detection performance.

Finally, as inferred by the blob detection rate after the stimulation of FEF in Experiment 2, the stimuli in the attended eye should dominate conscious perception in a majority of time during adaptation. As surmised on the basis of the SSVEP results of our previous work (Song, Lyu, Zhao, et al., 2023), there were probably stronger monocular neural responses to the inputs from the attended eye than those from the unattended eye in Experiment 2, too. Based on the homeostatic compensation theory (Lunghi et al., 2013; Turrigiano, 1999), this would result in a shift of ocular dominance towards the unattended eye following the adaptation. Nevertheless, we did not observe a significant change of ocular dominance after adaptation under the stimulation of FEF, which contradicts the prediction of homeostatic compensation theory. Therefore, the present findings provide direct causal evidence for the role of eye-based attention in modulating ocular dominance, and also indicate that this causal role of eye-based attention is distinct from the homeostatic compensation mechanism commonly used to explain the effect of short-term monocular deprivation (Bai et al., 2017; Lunghi et al., 2011; Lunghi et al., 2015; Lyu et al., 2020; Zhou et al., 2015; Zhou et al., 2013).

Materials and Methods

Experiment 1

Subjects

A total of twenty participants were recruited to take part in the screening process (see Stimuli and Procedure), 17 successfully passed the screening test and completed an fMRI experiment. Due to excessive head motion during fMRI scanning, the data of one participant were excluded from further analysis. Eventually, data from 16 subjects (12 females, aged from 20 to 28 years) were used. The sample size was predetermined based on the previous study (Song, Lyu, Zhao, et al., 2023). All participants were naive to the experimental hypotheses. They possessed normal or corrected-to-normal visual acuity, and provided informed consent in accordance with the Declaration of Helsinki. This study was approved by the Institutional Review Board of the Institute of Psychology, Chinese Academy of Sciences.

Apparatus

During the screening process, stimuli were presented on a gamma-corrected 27-inch ASUS VG279QM LCD monitor (Asus, Shanghai, China, 1920 × 1080 pixels’ resolution, 60 Hz refresh rate) with the mean luminance of 30.18 cd/ m2. Participants viewed the stimuli through a paper red-blue stereoscopic glasses at a distance of 100 cm. A chinrest was used to stabilize head position. The experimental stimuli were programmed using MATLAB (The MathWorks, Natick, USA) and Psychtoolbox (Brainard, 1997; Pelli, 1997). A binocular rivalry task was adopted to screen subjects for extreme ocular dominance.

During the fMRI scanning, the stimuli were projected onto a flat panel screen (49 × 36.75 cm, 1024 × 768 pixels’ resolution, 60 Hz refresh rate). The screen was gamma-corrected and placed in the front of the MR scanner using an LCD projector. Participants viewed the screen at the distance of 93 cm via the paper red-blue stereoscopic glasses. The experiment was conducted in a dimly lit room.

Stimuli and Procedure

Screen phase

The stimuli were two orthogonal sinusoidal gratings (±45° from vertical, 2° in diameter, 1.5 cpd) overlapped at the center of screen. We carefully adjusted the contrast of two gratings, so that when participants wore the red-blue glasses, they could only see the red grating through the red glass and the blue grating through the blue glass. The contrasts of red and blue gratings were 48% and 80%, respectively. In addition, a high-contrast checkerboard ‘‘frame” (size: 5°× 5°; thick: 0.25°) and a central white fixation point (diameter: 0.04°) were presented to facilitate binocular fusion (Figure 6).

Illustration of binocular rivalry stimuli in the screening process.

The binocular rivalry test consisted of sixteen 1-min trials. Each trial began with a 5-s blank interval, followed by the presentation of the rival gratings for 55 s. Participants were instructed to hold down one of the three keys (Right, Left, or Down Arrow) to report the grating they perceived (clockwise, counterclockwise, or mixed). The orientation of grating presented to each eye was constant within a trial, but was counter-balanced across trials.

fMRI scanning phase

During fMRI scanning, subjects viewed movie images either dichoptically or binocularly with the red-blue glasses. In the dichoptic condition, the movie stimuli were two overlapped red and blue movie episodes (Figure 2A). One was played normally and the other was the same movie but played in a backward sequence (Song, Lyu, Zhao, et al., 2023). The regular movie images were presented to the dominant eye and the backward ones to the non-dominant eye (for the definition of eye dominance, see Experimental Design below). In the binocular condition, the stimuli were regular movie images colored in purple to ensure visible inputs from both eyes (Figure 2B).

To create the colored movie stimuli, the original movie images were firstly transformed into grayscale. In theory, the red (or blue) movie images can then be generated by zeroing the values of Green and Blue (or Red and Green) channels. However, empirically, the contrast of the red movie images was reduced by applying a multiplication factor of 0.6 to the values of Red channel and setting the value of Blue channel to 80 to achieve perfect monocular stimulation in each eye. The red movie images and the blue movie images were then overlapped to generate the red-blue movie images used in the dichoptic condition (Figure 2D). The purple movie images were obtained by first generating the same red and blue regular movie images in a similar methodology as that employed for the aforementioned red-blue movie images, followed by their overlap. To facilitate binocular fusion, a gray fixation point (diameter: 0.33°) positioned at the center of the movie images and a high-contrast checkerboard ‘‘frame” (size: 17.73°× 23.85°; thick: 0.27°) surrounding the movie were presented during stimulation. The frame rate of movies was 30 fps.

A block design was adopted to study the neural mechanisms related to eye-based attention. A movie block lasted for 30 s, alternated with 12-18 s (12, 14, 16, or 18 s) blanks. Each run consisted of 8 movie blocks. The stimuli of each block could be either the dichoptic movie or the binocular movie, with the sequence counter-balanced.

The fMRI experiment included experimental runs and control runs. During an experimental run, participants were instructed to attentively watch the regular movie and comprehend the plot. To facilitate participants’ understanding of the plot, the sound track for the movie was played synchronized with the regular movie images. After the scanning of each run, a question on the movie plot would be presented on the screen. Subjects needed to answer the question via button press, allowing us to examine if they had well followed the plot during the scanning. In a control run, subjects performed a rapid serial visual presentation (RSVP) task while watching the movie stimuli (Figure 2C). A series of capital letters were presented to both eyes on the center of the screen. Each letter subtended 0.23°× 0.27°, and was presented for 153 ms. Participants’ task was to press the button whenever they found an ‘X’ in the letter stream. Within each block, the letter ‘X’ appeared for a total of three times. Participants received a feedback about their task performance after the completion of each run. To ensure that participants focused their attention on the RSVP task, the sound tracks were not played during the control runs.

Experimental Design

In the screening test, participants performed three sessions of binocular rivalry tests to measure perceptual ocular dominance. In the binocular rivalry tests, at any given time only the grating in one eye reached awareness and remained visible for a while before the grating in the other eye competed for awareness. The dominant eye was defined as the one that showed longer summed phase durations. Because perceptual ocular dominance fluctuated greatly in the first few trials of a day (Suzuki & Grabowecky, 2007), 5 warm-up trials were completed before the formal task session, the data of which were not analyzed. Perceptual ocular dominance in each test was evaluated by computing an ODI with the formula 𝑇𝐿/(𝑇𝐿 + 𝑇𝑅), where TL and TR represented the summed phase durations for perceiving the stimulus in the left eye and right eye, respectively. Extreme ocular dominance can potentially impede accurate measures of attentional effects. For instance, if the dominant eye is excessively strong, participants may effortlessly perceive the regular movie images for an extended duration, which makes voluntary eye-based attention dispensable. Therefore, participants with ODI values greater than 0.67 or less than 0.33 (3 participants in total) were not allowed to participate in the fMRI scanning.

Each participant completed 3 experimental runs and 3 control runs, with the order counter-balanced across participants. Each run consisted of the movie blocks of dichoptic and binocular conditions.

fMRI Data Acquisition and Preprocessing

fMRI data were collected by a Siemens 3T Magnetom Trio scanner with a 20-channel phased array head coil. T1 weighted anatomical images were acquired at the beginning of scanning (repetition time (TR) = 2600 ms, echo time (TE) = 3.02 ms, flip angle (FA) = 8°, field of view (FOV) = 256 mm, slices = 176, voxel size = 1.0 mm × 1.0 mm × 1.0 mm). Functional data were acquired with T2* weighted Echo-planar imaging (EPI) sequence (TR = 2000 ms, TE = 30 ms, FA = 90°, FOV = 200 mm, slices = 32, voxel size = 3.1 mm × 3.1 mm × 3.5 mm).

The first three volumes were discarded for magnetization equilibrium. The fMRI data were preprocessed using Analysis of Functional NeuroImages (AFNI) software (Cox, 1996). The preprocessing steps encompassed slice timing correction, motion correction, aligning the functional image with anatomy, spatial normalization (MNI152 space), spatial smoothing with 4 mm Gaussian kernel (full width at half maximum, FWHM), and scaling each voxel time series to have a mean of 100. Data from one subject was excluded from further analysis due to the head motion of more than 10% of volumes exceeding 0.3 mm (Silvers et al., 2017; Somerville et al., 2013).

fMRI Data Analysis

We performed single-subject general liner model (GLM) using the AFNI program “3dDeconvolve”. To create regressors for each experimental conditions (dichoptic and binocular), the boxcar function representing the duration of each block (30 s) was convolved with the canonical hemodynamic response function (HRF). Volumes were censored if the motion derivatives exceeded 0.3 mm. High-pass filtering and 3 motion parameters were modeled as non-interest regressors to account for drift and residual motion effects. In addition, single-subject contrasts between experiment conditions (dichoptic > binocular) were calculated.

Subsequently, we used AFNI program “3dttest++” to access group difference of the ‘dichoptic-binocular’ contrast between the experimental and control runs. To constrict the analysis on positive activations to movie stimuli, a mask was created specifically for the t-test analysis. This mask was generated by taking the union of all positively activated voxels across all participants for each experiment conditions. “ClustSim” was then applied for multiple comparison correction, yielding a minimum cluster size of 21 voxels (voxelwise p = .001; cluster threshold α = 0.05).

Experiment 2

Subjects

Twenty-one subjects were initially recruited to participate in the experiment. Eighteen of them successfully passed the screening and completed the formal experiment. The specific screening criteria can be found in the Experimental Design and Data Analysis section. Due to the failure of binocular fusion and the lack of concentration during the experiment, data of two subjects were excluded from the analysis. Accordingly, the data from 16 subjects (12 females, aged from 18 to 28 years) were analyzed, among which 5 subjects have participated in Experiment 1.

Apparatus

Visual stimuli of the behavioral task were displayed on a gamma-corrected 27-inch ASUS VG279QM LCD monitor with the mean luminance of 29.64 cd/ m2. The spatial resolution of monitor was 1920 × 1080 pixels, and the refresh rate was 60 Hz. Participants viewed visual stimuli through a mirror stereoscope with their eyes 100 cm away from the monitors. To stabilize their head position and minimize movement, participants were instructed to rest their heads on a chinrest. The behavioral tasks were performed in a dimly lit room.

Stimuli and Procedure

Binocular Rivalry Test

The binocular rivalry stimuli were two achromatic sine-wave grating disks that were oriented orthogonally (±45° from vertical, diameter: 1°, spatial frequency: 3 cpd, 80% Michelson contrast) and presented foveally to each eye, respectively. A central red fixation point (0.07° in diameter) and a high-contrast checkerboard ‘‘frame” (size: 2.5°×2.5°; thickness: 0.25°) were also presented to both eyes to facilitate stable binocular fusion (Figure 1A). The procedure of binocular rivalry test was consistent with that of Experiment 1.

Dichoptic-backward-movie Adaptation

We employed the “dichoptic-backward-movie” paradigm for the adaptation phase as utilized in the previous research (Song, Lyu, Zhao, et al., 2023). During the 50-min adaptation, participants were presented with two movie images dichoptically (Figure 1B), with one eye viewing the regular movie images while the other eye viewing the corresponding backward movie images. Note that the backward movie was formally identical to the regular one, except for the absence of a coherent plot. Participants were instructed to exert their utmost effort in comprehending the logic of the regular movie, while disregarding the superimposed backward movie. Therefore, top-down selective attention was supposed to be predominantly directed towards the regular movie. The eye that received the regular movie images was referred to as the “attended eye”, while the other eye as the “unattended eye”. The frame rate of movies was 30 fps.

Blob detection task

The blob detection task has been detailedly described in a prior research (Song, Lyu, Zhao, et al., 2023). During the adaptation phase, participants were required to watch movies while simultaneously detecting a reduction in color saturation within a blob region that was presented exclusively to one eye (Figure 1C). The locations of the blobs were predetermined by the experimenter, so that they always appeared on faces in the movie. Participants were required to promptly press the SPACE key upon detecting any portion of a character’s face turning gray.

To ensure that the primary focus remained on movie watching, the presentation of the blob was infrequent. To determine the timing of the presentation of the blobs, the adaptation period was divided into 20 segments, each lasting for 150 s. A blob would fade in within 0.2 s at a random moment in the middle 50-s of a segment on the movie of one eye. And after 5 s, it faded out within 0.2 s. There would be 10 blobs presented to each eye.

TMS

We utilized the Magstim Rapid2 stimulator equipped with a standard 70 mm figure-of-eight coil to deliver cTBS. The cTBS protocol consisted of 267 bursts, with each burst comprising 3 pulses at 30 Hz and repeating at 6 Hz. In total, there were 801 pulses delivered and the duration of a single cTBS train was 44 s (Cazzoli et al., 2009; Chaves et al., 2012; Dordevic et al., 2022; Nyffeler et al., 2008). To extend the duration of inhibitory effects induced by the cTBS, two cTBS trains with an interval of 15-min were applied over the same stimulation site. This stimulation protocol has been demonstrated to safely extend the duration of cTBS effects to over 2 hours (Corbetta & Shulman, 2002; Mayrhofer et al., 2019), thereby ensuring the sustained inhibitory effect throughout the adaptation process. The stimulation intensity was set at 80% of the individual’s resting motor threshold (RMT), which was defined as the minimum machine output required to elicit a visible twitch of the left index finger for 50% of pulses.

Given that the dorsal attentional network primarily consists of the FEF and the IPS (Corbetta & Shulman, 2002; Mayrhofer et al., 2019), with a functional right-hemisphere dominance (Duecker et al., 2013; Mayrhofer et al., 2019; Sack, 2010), we selected the right FEF and right IPS from the four clusters identified in Experiment 1 as the target regions for cTBS (Gallotto et al., 2022). The cTBS stimulation sites were localized based on the peak of the group effect observed in Experiment 1 (MNI coordinates: rFEF: 34.5, −1.5, 55.5; rIPS: 31.5, −73.5, 46.5. see Figure 4B). In addition, we applied cTBS on the vertex (MNI coordinates: 0, 0, 75) as a control condition (Wang et al., 2020). Coil position was guided by the Brainsight 2 neuro-navigation software based on individual T1-weighted images.

Experimental design

Before the formal experiment, all participants underwent a 3-7 days of training period (3 tests per day with a 10-minute rest in between) to ensure consistent performance on the binocular rivalry task (Bao et al., 2018). The task procedure of each day was similar to that of the screen test in Experiment 1. One subject withdrew from the study due to difficulties in achieving binocular fusion.

After the practice of binocular rivalry task, participants were also examined for their capacity in allocating eye-based attention (Neisser & Becklen, 1975). They were asked to watch the dichoptic movie while performing the blob detection task (Song, Lyu, Zhao, et al., 2023). Only those who showed superior blob detection performance for the attended eyes and subjectively reported paying more attention to the regular movie images were eligible to finish the formal experiment. All the subjects who participated in this stage successfully passed the screening. The brain structural images of subjects who had not previously participated in Experiment 1 were then acquired.

In the formal experiment, participants performed two binocular rivalry tests initially. The first served as a warm-up test that consisted of five trials and was not analyzed (Bai et al., 2017; Bao et al., 2018). The second comprised 16 trials that measured the perceptual ocular dominance prior to adaptation (i.e. pre-test). Following the pre-test, two cTBS trains with an interval of 15 min were applied over one stimulation site. Afterwards, participants completed a 50-min dichoptic-backward-movie adaption. During the adaptation phase, the main task was to comprehend the plot of the movie. Meanwhile, participants were required to detect the infrequent blob targets. The dichoptic movie was played with sound and the audio track always synchronized with the regular movie images to facilitate the allocation of eye-based attention. As shown in the previous work (Song, Lyu, Zhao, et al., 2023), presenting the regular movie images to the dominant eye can maximize the effect of eye dominance shift. Therefore, here the dominant eye was always selected as the attended eye. After the end of adaptation, a 16-trial binocular rivalry test was immediately conducted as the post-test (Figure 4A). The formal experiment consisted of three sessions, with cTBS delivered to only one of the three stimulation sites in each session. The order of stimulation site was counter-balanced across subjects. The sessions were conducted on different days to minimize any potential carryover effect.

Data Analysis

To quantify perceptual eye dominance, we calculated the ODI based on binocular rivalry dynamics. The ODI was derived using the following formula that yielded a value ranging from 0 (indicating complete dominance of the attended eye) to 1 (indicating complete dominance of the unattended eye). In the above formula, TAE and TUAE denote the cumulative phase durations of perceiving the stimulus in the attended eye and stimulus in the unattended eye, respectively. Participants with ODI values larger than 0.67 or smaller than 0.33 (2 participants in total) were not allowed to participate in the formal experiment.

Experiment 3

Subjects

Twenty subjects were recruited to participate in the experiment, of which two were excluded for extreme interocular imbalance and another two voluntarily withdrew during the screening phase. Accordingly, 16 subjects (14 females, aged from 19 to 26 years) completed all the experiments, one of which had participated in Experiment 1.

Stimuli and Procedure

The task procedure was similar to that in Experiment 2 except that participants watched identical regular movie images with both eyes and were no longer required to perform the blob detection task during adaptation.

eLife Assessment

This important study combines psychophysics, fMRI, and TMS to reveal a causal role of FEF in generating an attention-induced ocular dominance shift, with potential relevance for clinical applications. The evidence supporting the claims of the authors is solid, but the theoretical and mechanistic interpretation of results and experimental approaches need to be strengthened. The work will be of broad interest to perceptual and cognitive neuroscience.

Reviewer #1 (Public Review):

Summary:

Based on a "dichoptic-background-movie" paradigm that modulates ocular dominance, the present study combines fMRI and TMS to examine the role of the frontoparietal attentional network in ocular dominance shifts. The authors claimed a causal role of FEF in generating the attention-induced ocular dominance shift.

Strengths:

A combination of fMRI, TMS, and "dichoptic-background-movie" paradigm techniques is used to reveal the causal role of the frontoparietal attentional network in ocular dominance shifts. The conclusions of this paper are mostly well supported by data.

Weaknesses:

The relationship between eye dominance, eye-based attention shift, and cortical functions remains unclear and merits further delineation. The rationale of the experimental design related to the hemispheric asymmetry in the FEF and other regions should be clarified.

Theoretically, how the eye-related functions in this area could be achieved, and how it interacts with the ocular representation in V1 warrant further clarification.

Reviewer #2 (Public Review):

Summary

Song et al investigate the role of the frontal eye field (FEF) and the intraparietal sulcus (IPS) in mediating the shift in ocular dominance (OD) observed after a period of dichoptic stimulation during which attention is selectively directed to one eye. This manipulation has been previously found to transiently shift OD in favor of the unattended eye, similar to the effect of short-term monocular deprivation. To this aim, the authors combine psychophysics, fMRI, and transcranial magnetic stimulation (TMS). In the first experiment, the authors determine the regions of interest (ROIs) based on the responses recorded by fMRI during either dichoptic or binocular stimulation, showing selective recruitment of the right FEF and IPS during the dichoptic condition, in line with the involvement of eye-based attention. In a second experiment, the authors investigate the causal role of these two ROIs in mediating the OD shift observed after a period of dichoptic stimulation by selectively inhibiting with TMS (using continuous theta burst stimulation, cTBS), before the adaptation period (50 min exposure to dichoptic stimulation). They show that, when cTBS is delivered on the FEF, but not the IPS or the vertex, the shift in OD induced by dichoptic stimulation is reduced, indicating a causal involvement of the FEF in mediating this form of short-term plasticity. A third control experiment rules out the possibility that TMS interferes with the OD task (binocular rivalry), rather than with the plasticity mechanisms. From this evidence, the authors conclude that the FEF is one of the areas mediating the OD shift induced by eye-selective attention.

Strengths

1. The experimental paradigm is sound and the authors have thoroughly investigated the neural correlates of an interesting form of short-term visual plasticity combining different techniques in an intelligent way.

2. The results are solid and the appropriate controls have been performed to exclude potential confounds.

3. The results are very interesting, providing new evidence both about the neural correlates of eye-based attention and the involvement of extra-striate areas in mediating short-term OD plasticity in humans, with potential relevance for clinical applications (especially in the field of amblyopia).

Weaknesses

1. Ethics: more details about the ethics need to be included in the manuscript. It is only mentioned for experiment 1 that participants "provided informed consent in accordance with the Declaration of Helsinki. This study was approved by the Institutional Review Board of the Institute of Psychology, Chinese Academy of Sciences". (Which version of the Declaration of Helsinki? The latest version requires the pre-registration of the study. The code of the approved protocol together with the code and date of the approval should be provided.) There is no mention of informed consent procedures or ethics approval for the TMS experiments. This is a huge concern, especially for brain stimulation experiments!

2. Statistics: the methods section should include a sub-section describing in detail all the statistical analyses performed for the study. Moreover, in the results section, statistical details should be added to support the fMRI results. In the current version of the manuscript, the claims are not supported by statistical evidence.

3. Interpretation of the results: the TMS results are very interesting and convincing regarding the involvement of the FEF in the build-up of the OD shift induced by dichoptic stimulation, however, I am not sure that the authors can claim that this effect is related to eye-based attention, as cTBS has no effect on the blob detection task during dichoptic stimulation. If the FEF were causally involved in eye-based attention, one would expect a change in performance in this task during dichoptic stimulation, perhaps a similar performance for the unattended and attended eye. The authors speculate that the sound could have an additional role in driving eye-based attention, which might explain the lack of effect for the blob discrimination task, however, this hypothesis has not been tested.

4. Writing: in general, the manuscript is well written, but clarity should be improved in certain sections.

a. fMRI results: the first sentence is difficult to understand at first read, but it is crucial to understand the results, please reformulate and clarify.

b. Experiment 3: the rationale for experiment one should be straightforward, without a long premise explaining why it would not be necessary.

c. Discussion: the language is a bit familiar here and there, a more straightforward style should be preferred (one example: p.19 second paragraph).

5. Minor: the authors might consider using the term "participant" or "observer" instead of "subject" when referring to the volunteers who participated in the study.

Reviewer #3 (Public Review):

Summary:

This study studied the neural mechanisms underlying the shift of ocular dominance induced by "dichoptic-backward-movie" adaptation. The study is self-consistent.

Strengths:

The experimental design is solid and progressive (relationship among three studies), and all of the raised research questions were well answered.

The logic behind the neural mechanisms is solid.

The findings regarding the cTMS (especially the position/site can be useful for future medical implications).

Weaknesses:

Why does the "dichoptic-backward-movie" adaptation matter? This part is severely missing. This kind of adaptation is neither intuitive like the classical (Gbison) visual adaptation, nor practical as adaptation as a research paradigm as well as the fundamental neural mechanism. If this part is not clearly stated and discussed, this study is just self-consistent in terms of its own research question. There are tons of "cool" phenomena in which the neural mechanisms are apparent as "FEF controls vision-attention" but never tested using TMS & fMRI, but we all know that this kind of research is just of incremental implications.

Author Response

The following is the authors’ response to the original reviews.

eLife assessment

This important study combines psychophysics, fMRI, and TMS to reveal a causal role of FEF in generating an attention-induced ocular dominance shift, with potential relevance for clinical applications. The evidence supporting the claims of the authors is solid, but the theoretical and mechanistic interpretation of results and experimental approaches need to be strengthened. The work will be of broad interest to perceptual and cognitive neuroscience.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

Based on a "dichoptic-background-movie" paradigm that modulates ocular dominance, the present study combines fMRI and TMS to examine the role of the frontoparietal attentional network in ocular dominance shifts. The authors claimed a causal role of FEF in generating the attention-induced ocular dominance shift.

Strengths:

A combination of fMRI, TMS, and "dichoptic-background-movie" paradigm techniques is used to reveal the causal role of the frontoparietal attentional network in ocular dominance shifts. The conclusions of this paper are mostly well supported by data.

Weaknesses:

(1) The relationship between eye dominance, eye-based attention shift, and cortical functions remains unclear and merits further delineation. The rationale of the experimental design related to the hemispheric asymmetry in the FEF and other regions should be clarified.

Thanks for the reviewer’s comments! We have further clarified the relationship between eye dominance shift, eye-based attention, and cortical functions in the Introduction and Discussion. In the Introduction, we introduce the modulating effects of eye-based attention on eye dominance. On one hand, eye-based attention can enhance eye dominance of the attended eye in real time (see page 3 first paragraph or below):

”For instance, presenting top-down attentional cues to one eye can intensify the competition strength of input signals in the attended eye during binocular rivalry (Choe & Kim, 2022; Zhang et al., 2012) and shift the eye balance towards the attended eye (Wong et al., 2021).”

On the other hand, prolonged eye-based attention can induce a shift of eye dominance to the unattended eye (see page 3 second paragraph or below):

“In Song et al. (2023)’s “dichoptic-backward-movie” adaptation paradigm (see Figure 1B), participants are presented with regular movie images in one eye (i.e., attended eye) while the other eye (i.e., unattended eye) received the backward movie images of the same episode. They were also instructed to try their best to follow the logic of the regular movie and ignore the superimposed backward movie. Therefore, the goal-directed eye-based attention was predominantly focused on the attended eye. Song et al. (2023) found that the predominance of the unattended eye in binocular rivalry increased after one hour of adaptation to the “dichoptic-backward-movie”, indicating a shift of perceptual ocular dominance towards the unattended eye. Since the overall energy of visual input from the two eyes was balanced throughout the adaptation period, the change of ocular dominance after adaptation is thought to result from unbalanced eye-based attention rather than unbalanced input energy as in typical short-term monocular deprivation (Bai et al., 2017; Lunghi et al., 2011; Zhou et al., 2014).”

Moreover, we discussed how FEF regulates attention-induced ocular dominance shift (see page 21 second paragraph to page 23 first paragraph or below, which also respond to this reviewer’s comment of Weakness #2):

“Then how does FEF regulate the attention-induced ocular dominance shift? Our previous work has found that the aftereffect (for simplicity, hereafter we use aftereffect to denote the attention-induced ocular dominance shift) can be produced only when the adapting stimuli involve adequate interocular competition, and is measurable only when the testing stimuli are not binocularly fused (Song et al., 2023). Given the indispensability of interocular competition, we explained those findings in the framework of the ocular-opponency-neuron model of binocular rivalry (Said & Heeger, 2013). The model suggests that there are some opponency neurons which receive excitatory inputs from monocular neurons for one eye and inhibitory inputs from monocular neurons for the other eye (e.g. AE-UAE opponency neurons receive excitatory inputs from the attended eye (AE) and inhibitory inputs from the unattended eye (UAE)). Then a difference signal is computed so that the opponency neurons fire if the excitatory inputs surpass the inhibitory inputs. Upon activation, the opponency neurons will in turn suppress the monocular neurons which send inhibitory signals to them.

Based on this model, we proposed an ocular-opponency-neuron adaptation account to explain the aftereffect, and pointed out that the attentional system likely modulated the AE-UAE ocular opponency neurons (Song et al., 2023). So why would FEF modulate the AE-UAE opponency neurons? The reason may be two fold. Firstly, understanding the logic during the dichoptic-backward-movie viewing may require filtering out the distracting information (from the unattended eye) and sustaining attention (to the attended eye), which is exactly the role of FEF (Esterman et al., 2015; Lega et al., 2019).

Secondly, due to the special characteristics of binocular vision system, filtering the distracting input from the unattended eye may have to rely on the interocular suppression mechanism. According to the ocular-opponency-neuron model, this is achieved by the firing of the AE-UAE opponency neurons that send inhibitory signals to the UAE monocular neurons.

As mentioned previously, the firing of the AE-UAE opponency neurons requires stronger activity for the AE monocular neurons than for the UAE monocular neurons. This is confirmed by the results shown in Figure 8 of Song et al. (2023) that monocular response for the attended eye during the entire adaptation phase was slightly stronger than that for the unattended eye. Accordingly, during adaptation the AE-UAE opponency neurons were able to activate for a longer period thus adapted to a larger extent than the UAE-AE opponency neurons. This would cause the monocular neurons for the unattended eye to receive less inhibition from the AE-UAE opponency neurons in the post-test as compared with the pre-test, leading to a shift of ocular dominance towards the unattended eye. In this vein, the magnitude of this aftereffect should be proportional to the extent of adaptation of the AE-UAE relative to UAE-AE opponency neurons. Attentional enhancement on the AE-UAE opponency neurons is believed to strengthen this aftereffect, as it has been found that attention can enhance adaptation (Dong et al., 2016; Rezec et al., 2004). Inhibition of FEF likely led such attentional modulation to be much less effective. Consequently, the AE-UAE opponency neurons might not have the chance to adapt to a sufficiently larger extent than the UAE-AE opponency neurons, leading to a statistically non-detectable aftereffect in Experiment 2. Therefore, the results of Experiments 2-4 in the present study suggest that within the context of the ocular-opponency-neuron adaptation account, FEF might be the core area to fulfill the attentional modulations on the AE-UAE opponency neurons.”

We used the experimental design with hemispheric asymmetry in the FEF and other regions for two reasons. First, many studies have shown that the dorsal attentional network has a functional right-hemisphere dominance (Duecker et al., 2013; Mayrhofer et al., 2019; Sack, 2010). This was also indicated by the results of Experiment 1 (Figure 3). Second, we found that a recent research applying TMS to FEF and IPS stimulated only the right hemisphere (Gallotto et al., 2022). Therefore, we selected the right FEF and right IPS as the target regions for cTBS. In the Methods section of Experiment 2, we have elucidated the reasons for the selection of cTBS target regions (see page 35, first paragraph or below):

“Given that the dorsal attentional network primarily consists of the FEF and the IPS (Corbetta & Shulman, 2002; Mayrhofer et al., 2019), with a functional right-hemisphere dominance (Duecker et al., 2013; Mayrhofer et al., 2019; Sack, 2010), we selected the right FEF and right IPS from the four clusters identified in Experiment 1 as the target regions for cTBS (Gallotto et al., 2022).”

(2) Theoretically, how the eye-related functions in this area could be achieved, and how it interacts with the ocular representation in V1 warrant further clarification.

Thanks for the reviewer’s comment! In the revised manuscript, we have discussed how FEF regulates attention-induced ocular dominance shift (see page 21 second paragraph to page 23 first paragraph or the quoted paragraphs under this reviewer’s first Public comment).

Reviewer #2 (Public Review):

Summary

Song et al investigate the role of the frontal eye field (FEF) and the intraparietal sulcus (IPS) in mediating the shift in ocular dominance (OD) observed after a period of dichoptic stimulation during which attention is selectively directed to one eye. This manipulation has been previously found to transiently shift OD in favor of the unattended eye, similar to the effect of short-term monocular deprivation. To this aim, the authors combine psychophysics, fMRI, and transcranial magnetic stimulation (TMS). In the first experiment, the authors determine the regions of interest (ROIs) based on the responses recorded by fMRI during either dichoptic or binocular stimulation, showing selective recruitment of the right FEF and IPS during the dichoptic condition, in line with the involvement of eye-based attention. In a second experiment, the authors investigate the causal role of these two ROIs in mediating the OD shift observed after a period of dichoptic stimulation by selectively inhibiting with TMS (using continuous theta burst stimulation, cTBS), before the adaptation period (50 min exposure to dichoptic stimulation). They show that, when cTBS is delivered on the FEF, but not the IPS or the vertex, the shift in OD induced by dichoptic stimulation is reduced, indicating a causal involvement of the FEF in mediating this form of short-term plasticity. A third control experiment rules out the possibility that TMS interferes with the OD task (binocular rivalry), rather than with the plasticity mechanisms. From this evidence, the authors conclude that the FEF is one of the areas mediating the OD shift induced by eye-selective attention.

Strengths

(1) The experimental paradigm is sound and the authors have thoroughly investigated the neural correlates of an interesting form of short-term visual plasticity combining different techniques in an intelligent way.

(2) The results are solid and the appropriate controls have been performed to exclude potential confounds.

(3) The results are very interesting, providing new evidence both about the neural correlates of eye-based attention and the involvement of extra-striate areas in mediating short-term OD plasticity in humans, with potential relevance for clinical applications (especially in the field of amblyopia).

Weaknesses

(1) Ethics: more details about the ethics need to be included in the manuscript. It is only mentioned for experiment 1 that participants "provided informed consent in accordance with the Declaration of Helsinki. This study was approved by the Institutional Review Board of the Institute of Psychology, Chinese Academy of Sciences". (Which version of the Declaration of Helsinki? The latest version requires the pre-registration of the study. The code of the approved protocol together with the code and date of the approval should be provided.) There is no mention of informed consent procedures or ethics approval for the TMS experiments. This is a huge concern, especially for brain stimulation experiments!

Response: Thanks for the reviewer’s comment! In the revised manuscript, we have provided the code of the approved protocol and date of the approval (see page 25 second paragraph or below):

“This study was approved (H21058, 11/01/2021) by the Institutional Review Board of the Institute of Psychology, Chinese Academy of Sciences.”

Indeed, ethics approval and informed consent were obtained for each experiment. To avoid duplication in the text, we only presented the ethics instructions in the Methods section of Experiment 1. We have now clarified in that section that all the experiments in this study were approved by the IRB in our Institute.

(2) Statistics: the methods section should include a sub-section describing in detail all the statistical analyses performed for the study. Moreover, in the results section, statistical details should be added to support the fMRI results. In the current version of the manuscript, the claims are not supported by statistical evidence.

Response: Thanks for the reviewer’s suggestion! In the Methods section of revised manuscript, we have added a section to describe the detailed statistical analyses for each experiment (see page 37 last paragraph for Experiment 2 and page 38 last paragraph for Experiment 3 or below):

“Statistical analyses were performed using MATLAB. A 3 (stimulation site: Vertex, FEF, IPS) × 2 (test phase: pre-test and post-test) repeated measures ANOVA was used to investigate the effect of cTBS delivery on ocular dominance shift. Moreover, for the blob detection test, the target detection rate of each experimental condition was calculated by dividing the summed number of detected blob targets by the total number of blob targets. Then, a 2 (eye: attended eye, unattended eye) × 3 (stimulation site: Vertex, FEF, IPS) repeated measures ANOVA on the detection performance was performed. Post-hoc tests were conducted using paired t-tests (2-tailed significance level at α = 0.05), and the resulting p-values were corrected for multiple comparisons using the false discovery rate (FDR) method (Benjamini & Hochberg, 1995).”

“In addition to the data analysis in Experiment 2, we complemented the standard inferential approach with the Bayes factor (van den Bergh et al., 2023; van Doorn et al., 2021; Wagenmakers et al., 2018), which allows quantifying the relative evidence that the data provide for the alternative (H1) or null hypothesis (H0). We conducted the Bayesian repeated measures ANOVA using JASP with default priors and computed inclusion Bayes factors (BFincl) which suggest the evidence for the inclusion of a particular effect calculated across matched models. A BF greater than 1 provides support for the alternative hypothesis. Specifically, a BF between 1 and 3 indicates weak evidence, a BF between 3 and 10 indicates moderate evidence, and a BF greater than 10 indicates strong evidence (van Doorn et al., 2021). In contrast, a BF below 1 provides evidence in favor of the null hypothesis.”

Furthermore, in the Results section of revised manuscript, we have added the statistical details to support the fMRI results (see page 9 last paragraph or below):

“To seek these brain regions, we used the AFNI program “3dttest++” to access the difference of ‘dichoptic-binocular’ contrast between the experimental and control runs. The AFNI program “ClustSim” was then applied for multiple comparison correction, yielding a minimum significant cluster size of 21 voxels (voxel wise p = .001; cluster threshold α = 0.05). We found 4 clusters showing stronger responses to the dichoptic movies than to the binocular movies especially in the experimental runs.”

(3) Interpretation of the results: the TMS results are very interesting and convincing regarding the involvement of the FEF in the build-up of the OD shift induced by dichoptic stimulation, however, I am not sure that the authors can claim that this effect is related to eye-based attention, as cTBS has no effect on the blob detection task during dichoptic stimulation. If the FEF were causally involved in eye-based attention, one would expect a change in performance in this task during dichoptic stimulation, perhaps a similar performance for the unattended and attended eye. The authors speculate that the sound could have an additional role in driving eye-based attention, which might explain the lack of effect for the blob discrimination task, however, this hypothesis has not been tested.

Response: Thanks for the reviewer’s comment! Following this reviewer’s insightful suggestion, we have conducted a new experiment to examine the effect of sound on blob detection task (see Experiment 4 in the revised manuscript). The procedure was similar to that of Experiment 2 except that the sound was no longer presented during the dichoptic-backward-movie adaptation. The results showed that the interocular difference of blob detection rate after sound elimination remained unaffected by the cTBS, which disagreed with our explanation in the previous version of manuscript. Based on the new data, we now question the validity to use the blob detection rate to precisely quantify eye-based attention, and have tried to explain why the blob detection results do not contradict with our account for the function role of FEF in modulating the aftereffect in the Discussion of the revised manuscript (see page 23 second paragraph to page 24 first paragraph or below):

“An unresolved issue is why inhibiting the cortical function of FEF did not impair the performance of blob detection task. One potential explanation is that the synchronized audio in Experiment 2 might help increase the length of time that the regular movie dominated awareness. However, the results of Experiment 4 did not support this explanation, in which the performance of blob detection survived from the inhibition of FEF even when silent movies were presented. Although this issue remains to be explored in future work, it does not contradict with our notion of FEF modulating AE-UAE opponency neurons. It should be noted that our notion merely states that FEF is the core area for attentional modulations on activities of AE-UAE opponency neurons. No other role of FEF during the adaptation is assumed here (e.g. boosting monocular responses or increasing conscious level of stimuli in the attended eye). In contrast, according to the most original definition, the blob detection performance serves as an estimation of visibility (or consciousness level) of the stimuli input from each eye, despite the initial goal of adopting this task is to precisely quantify eye-based attention (which might be impractical). Thus, according to our notion, inhibition of FEF does not necessarily lead to deteriorate performance of blob detection. Furthermore, our findings consistently indicated that the visibility of stimuli in the attended eye was markedly superior to that of stimuli in the unattended eye, yet the discrepancy in the SSVEP monocular responses between the two eyes was minimal though it had reached statistical significance (Song et al., 2023). Therefore, blob detection performance in our work may only faithfully reflect the conscious level in each monocular pathway, but it is probably not an appropriate index tightly associated with the attentional modulations on monocular responses in early visual areas. Indeed, previous work has argued that attention but not awareness modulates neural activities in V1 during interocular competition (Watanabe et al., 2011), but see (Yuval-Greenberg & Heeger, 2013). We have noticed and discussed the counterintuitive results of blob detection performance in our previous work (Song et al., 2023). Here, with the new counterintuitive finding that inhibition of FEF did not impair the performance of blob detection, we suspect that blob detection performance in the “dichoptic-backward-movie” adaptation paradigm may not be an ideal index that can be used to accurately quantify eye-based attention.

(4) Writing: in general, the manuscript is well written, but clarity should be improved in certain sections.

(a) fMRI results: the first sentence is difficult to understand at first read, but it is crucial to understand the results, please reformulate and clarify.

Response: Thanks for the reviewer’s suggestion! In the revised manuscript, we have reformulated this sentence (see page 9 last paragraph or below):

“It was only in the dichoptic condition of experimental runs that participants had to selectively pay more attention to one eye (i.e., eye-based attention). Therefore, we speculate that if certain brain regions exhibit greater activities in the dichoptic condition as compared to the binocular condition in the experimental runs but not in the control runs, the activation of these brain regions could be attributable to eye-based attention.”

(b) Experiment 3: the rationale for experiment one should be straightforward, without a long premise explaining why it would not be necessary.

Response: Thanks for the reviewer’s suggestion! In the revised manuscript, we have streamlined the lengthy premise explaining to make the rationale of Experiment 3 more straightforward (see page 15 last two paragraphs or below):

“The results of Experiment 2 support the notion that eye-based attention was the cause for attention-induced ocular dominance plasticity. However, an alternative account is that the significant two-way interaction between test phase and stimulation site did not stem from any persistent malfunction of FEF in modulating ocular dominance, but rather it was due to some abnormality of binocular rivalry measures in the post-test that occurred after stimulation at the FEF only (and not at the other two brain sites). For instance, stimulation at the FEF might simply reduce the ODI measured in the binocular rivalry post-test.

Therefore, we conducted Experiment 3 to examine how suppression of the three target sites would impact binocular rivalry performance, in case that any unknown confounding factors, which were unrelated to adaptation but related to binocular rivalry measures, contributed to the results.”

(c) Discussion: the language is a bit familiar here and there, a more straightforward style should be preferred (one example: p.19 second paragraph).

Response: Thanks for the reviewer’s suggestion! We have carefully revised the language in the discussion. The discussion following the example paragraph has been largely rewritten.

(5) Minor: the authors might consider using the term "participant" or "observer" instead of "subject" when referring to the volunteers who participated in the study.

Response: Thanks for the reviewer’s suggestion! In the revised manuscript, we have replaced the term “subject” with “participant”.

Reviewer #3 (Public Review):

Summary:

This study studied the neural mechanisms underlying the shift of ocular dominance induced by "dichoptic-backward-movie" adaptation. The study is self-consistent.

Strengths:

The experimental design is solid and progressive (relationship among three studies), and all of the raised research questions were well answered.

The logic behind the neural mechanisms is solid.

The findings regarding the cTMS (especially the position/site can be useful for future medical implications).

Weaknesses:

Why does the "dichoptic-backward-movie" adaptation matter? This part is severely missing. This kind of adaptation is neither intuitive like the classical (Gbison) visual adaptation, nor practical as adaptation as a research paradigm as well as the fundamental neural mechanism. If this part is not clearly stated and discussed, this study is just self-consistent in terms of its own research question. There are tons of "cool" phenomena in which the neural mechanisms are apparent as "FEF controls vision-attention" but never tested using TMS & fMRI, but we all know that this kind of research is just of incremental implications.

Response: Thanks for the reviewer’s comment! We designed the "dichoptic-backward-movie" adaptation to study the perceptual consequence and mechanisms of sustained attention to a monocular pathway. Since the overall visual input to both eyes during adaptation were identical, any effect (i.e. the change of ocular dominance in our study) after adaptation can be easily ascribed to unbalanced eye-based attention between the two eyes rather than unbalanced input energy across the eyes. In typical short-term monocular deprivation, input signal from one eye is blocked. Accordingly, attention is undoubtedly distributed to the non-deprived eye. The fact that in a short-term monocular deprivation paradigm the deprived eye is also the unattended eye prevents researchers from ascertaining whether unbalanced eye-based attentional allocation contributes to the shift of ocular dominance just like unbalanced visual input across the two eyes. That is why the “dichoptic-backward-movie” adaptation was adopted in the present study. This new paradigm balances the input energy across the eyes but leaves attention unbalanced across the eyes. In the revised manuscript, we have added the description of the “dichoptic-backward-movie” adaptation (see page 3 last paragraph and page 4 first paragraph or below). Hope this complementary information improves the clarity.

“In Song et al. (2023)’s “dichoptic-backward-movie” adaptation paradigm (see Figure 1B), participants are presented with regular movie images in one eye (i.e., attended eye) while the other eye (i.e., unattended eye) received the backward movie images of the same episode. They were also instructed to try their best to follow the logic of the regular movie and ignore the superimposed backward movie. Therefore, the goal-directed eye-based attention was predominantly focused on the attended eye. Song et al. (2023) found that the predominance of the unattended eye in binocular rivalry increased after one hour of adaptation to the “dichoptic-backward-movie”, indicating a shift of perceptual ocular dominance towards the unattended eye. Since the overall energy of visual input from the two eyes was balanced throughout the adaptation period, the change of ocular dominance after adaptation is thought to result from unbalanced eye-based attention rather than unbalanced input energy as in typical short-term monocular deprivation (Bai et al., 2017; Lunghi et al., 2011; Zhou et al., 2014).” In short-term monocular deprivation, input signal from one eye is blocked. Accordingly, attention is biased towards the non-deprived eye. However, it is difficult to tease apart the potential contribution of unbalanced eye-based attention from the consequence of the unbalanced input energy, as the deprived eye is also the unattended eye. Therefore, the advantage of the “dichoptic-backward-movie” adaptation paradigm is to balance the input energy across the eyes but leave attention unbalanced across the eyes.

Our previous work (Song et al., 2023) has shown that eye-based attention plays a role in the formation of ocular dominance shift following adaptation to dichoptic backward movie. However, because the “dichoptic-backward-movie” adaptation paradigm is new, to our knowledge, no literature has ever discovered the brain areas that are responsible for eye-based attention. Our fMRI experiment for the first time resolves this issue, which, we believe, is one of the novelties of the present study. Attention is a pretty general definition of our ability to select limited information for preferential or privileged processing, yet it includes numerous aspects (e.g. spatial attention for spatial locations, feature-based attention for visual features, object-based attention for objects, social attention for social cues, and eye-based attention for monocular pathways etc). Are we 100% sure that the same brain network always underlies every aspect of attention including eye-based attention? No test, no answer. Maybe the answer is Yes, but we are not aware of any evidence for that from literature. It is not unlikely that attention is like an elephant while researchers are like blind people touching the elephant from different angles. Even if all previous researchers have touched the side of the elephant and state that an elephant is no different from a wall, as long as one researcher grabs the elephant’s tail, the “wall” knowledge will be falsified. From this perspective of the essence of science (falsifiable), we have the confidence to say that our fMRI experiment on eye-based attention is novel, because to our knowledge our experiment is the first one to explore the issue. On the basis of the fMRI experiment (otherwise we would have no idea on which precise brain site to apply the cTBS), we could successfully complete the subsequent TMS experiments.

Of course, if the reviewer can kindly point out any previous neuroimaging work we missed that has already disclosed the neural mechanisms underlying human’s eye-based attention, we would truly appreciate the reviewer very much. But even so, we would like to emphasize that the purpose of the current study was actually not to use TMS & fMRI to confirm that “FEF controls visual attention”. As we mentioned in the Abstract and expanded the introduction in the last two paragraphs of Introduction, the goal of the TMS experiments is to examine the causal role of eye-based attention in producing the aftereffect of “dichoptic-backward-movie” adaptation. This research question is also new, thus we do not think the TMS experiments are incremental, either. Our findings provided direct causal evidence for the effect of FEF on modulating ocular dominance through eye-based attention. Please see the last two sentences in the first paragraph on page 20 in the revised manuscript or below,

“Interestingly, in our Experiment 2 this aftereffect was significantly attenuated after we temporarily inhibited the cortical function of FEF via cTBS. This finding indicates the crucial role of FEF in the formation of attention-induced ocular dominance shift.”

as well as the last sentence of the Abstract,

“…and in this network, FEF plays a crucial causal role in generating the attention-induced ocular dominance shift.”

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

(1) The hemispheric asymmetry in the eye-based attention-related cortex should be further examined and discussed. For example, IPS in both hemispheres was identified in the fMRI experiment. It is not clear why only the right IPS was stimulated in the TMS experiment.

Response: Thanks for the comment. We have elucidated the reasons for the experimental design with hemispheric asymmetry in FEF and IPS. Please see our response to the Weakness #1 raised by Reviewer #1 in the Public Review section.

(2) It is known that the frontoparietal cortex plays a role in the contralateral shift of attentional allocation. Meanwhile, the latest stage of ocular-specific representation is V1. The authors should discuss how the eye-related function can be achieved in FEF.

Response: Thanks for the comment. we have discussed how FEF regulates attention-induced ocular dominance shift (see page 21 second paragraph to page 23 first paragraph in the revised manuscript, and our response to the Weakness #2 raised by Reviewer #1 in the Public Review section).

(3) To further validate the role of FEF in eye-related attention shifts, the authors may consider using the traditional monocular deprivation paradigm with fMRI and TMS. It would be valuable to compare the neural mechanisms related to the classical monocular deprivation paradigm with the current findings.

Response: Thanks for the reviewer’s suggestion! That is indeed an interesting research topic that we are currently exploring. The current study investigated the attention-induced ocular dominance shift with the “dichoptic-backward-movie-adaptation” paradigm. This paradigm is substantially different from traditional short-term monocular deprivation. In our Neuroscience Bulletin paper (Song et al. 2023), we discuss the reason as follows.

“An alternative account of our results is the homeostatic plasticity mechanism. The function of this mechanism is to stabilize neuronal activity and prevent the neuronal system from becoming hyperactive or hypoactive. For this goal, the mechanism moves the neuronal system back toward its baseline after a perturbation [51, 52]. In our case, the aftereffect can be explained such that the visual system boosts the signals from the unattended eye to maintain the balance of the network’s excitability. However, this account cannot easily explain why the change of neural ocular dominance led by prolonged eye-based attention was observed here using the binocular rivalry testing stimuli, but absent in the previous research using the binocularly fused stimuli [11]. In contrast, a recent SSVEP study also using the binocularly fused stimuli has successfully revealed a shift of neural ocular dominance after two hours of monocular deprivation [31], which is in line with the homeostatic plasticity account. Therefore, the mechanisms underlying the “dichoptic-backward-movie” adaptation and monocular deprivation are probably not fully overlapped with each other; and the binocular rivalry mechanism described in the ocular-opponency-neuron model seems to be more preferable than the homeostatic plasticity mechanism in accounting for the present findings.”

Therefore, before asking whether FEF plays a role in the attention-induced ocular dominance shift in a traditional monocular deprivation paradigm, one should probably first examine whether attention also plays a role in traditional monocular deprivation, and whether the ocular-opponency-neuron adaptation account can also be used to explain the traditional monocular deprivation effect. Our newly accepted paper “Negligible contribution of adaptation of ocular opponency neurons to the effect of short-term monocular deprivation” (https://www.frontiersin.org/articles/10.3389/fpsyg.2023.1282113/full) gives a generally negative answer to the second question. And as to the first question, we have one manuscript under review and another ongoing study. In other words, to get a satisfactory answer to this particular comment of this reviewer, we need to first obtain clear answers to the two above questions. We think this is far beyond the scope of one single manuscript.

(4) The authors only presented regular movies to the dominant eye to maximize the ocular dominance shift. This critical information of design should be clarified, not only in the method section.

Response: Thanks for the reviewer’s suggestion! In the Results section of Experiment 2, we have added a description of this critical information of design (see page 11 last paragraph to page 12 first paragraph or below):

“Then, participants adapted to the “dichoptic-backward-movie” in which regular movie images were presented to the dominant eye to maximize the effect of eye dominance shift (Song et al., 2023). Meanwhile they were asked to detect some infrequent blob targets presented on the movie images in one eye at the same time.”

(5) The frame rate of the movie is 30 fps, which is much lower than a typical 60 fps visual presentation, does this have an effect on the adaptation outcome?

Response: To our best of knowledge, there is no evidence that the frame rate of the movie influences the aftereffect of attention-induced ocular dominance shift. In our previous research, the frame rate of the movie during adaptation was 25 fps, which still produced a stable adaptation aftereffect (Song et al., 2023). And the frame rate of the movie was 30 fps in our monocular deprivation work (Lyu et al., 2020), which showed a similar monocular deprivation effect we previously observed in an altered reality study (Bai et al., 2017). The frame rate of the altered-reality video in Bai et al.’s (2017) work was 60 fps. All these clues suggest that the frame rate does not have an effect on the adaptation outcome.

(6) Figure 5: The ODSE derived from ODI in Experiment 3 should also be illustrated, for a better comparison with results from Experiment 2.

Response: Thanks for the reviewer’s suggestion! In the revised manuscript, we have added the results of ODSE in Experiment 3 to Figure 5 (see page 15 or below):

Figure 5.

The results of (A) the ocular dominance index (ODI), (B) the ocular dominance shift effects (ODSE) in Experiment 2, (C) the ODI and (D) the ODSE in Experiment 3. The bars show the grand average data for each condition. The individual data are plotted with gray lines or dots. The dashed gray line represents the absolute balance point for the two eyes (ODI = 0.5). Error bars indicate standard errors of means. * p < .05; ** p < .01; n.s. p > .05.

(7) Spelling issues: "i.e." → "i.e.,"

Response: Thanks for the reviewer’s suggestion! In the revised manuscript, we have changed “i.e.” to “i.e.,”.

Reviewer #2 (Recommendations For The Authors):

Linked to weakness 3: Ideally, a control experiment with cTBS and dichoptic stimulation without sound but with the blob discrimination task should be performed to be able to make important claims about the neural mechanisms involved in eye-based attention.

Response: Thanks for the comment. We have performed a new experiment as the reviewer suggested. Please see our response to the Weakness #3 raised by Reviewer #2 in the Public Review section.

Reviewer #3 (Recommendations For The Authors):

(1) The neural mechanisms are so apparent. We all know the FEF\IPS\SC matter in vision and attention and gaze. This is not groundbreaking.

Response: As we addressed in our response to Reviewer #3’s public comment, the current study aimed at investigating the causal mechanism for eye-based attentional modulation of ocular dominance plasticity rather than simply the role of FEF\IPS\SC in visual attention. Moreover, eye-based attention is a less investigated aspect of visual attention. The neural mechanism underlying eye-based attention is still largely unknown, and seeking the brain areas for controlling eye-based attention is the necessary preparation work for applying the cTBS. We have responded in detail to Reviewer #3’s public comment why we think both the fMRI and TMS experiments are novel to the field, which we will not reiterate it here to avoid redundancy.

(2) Why does the "dichoptic-backward-movie" adaptation matter? Is playing a backward movie to one eye realistic? Does that follow the efficient coding? Is that a mere consequence of information theory?

Response: Thanks for the comments. We have added the description of the “dichoptic-backward-movie” adaptation paradigm in the revised manuscript (see page 3 last paragraph and page 4 first paragraph or our response to this reviewer’s Public comment).

Is it realistic to play backward movie to one eye? We feel this question is somehow ambiguous to us. If the reviewer means the technical operability for such stimulus presentation, we can assure it since we have used this paradigm in both the current and previously published studies. To be more specific, we made the video stimuli in advance. The left half of the video was the regular movie and the right half was the backward version of the same movie (or vice versa). When viewing such video stimuli through stereoscopes, participants could only see the left half of the video with the left eye and the right half of the video with the right eye. In other words, the regular movie and backward movie were viewed dichoptically. Alternatively, if the reviewer means that such dichoptic presentation rarely happens in real world thus not realistic, we agree with the reviewer on one hand. On the other hand, we have explained on page 3 last paragraph and page 4 first paragraph why it is a particular useful paradigm for the main purpose of the present study. Let us make a similar example. The phenomenon of binocular rivalry rarely happens in everyday life. So people may say binocular rivalry is not realistic. However, our visual system does have the ability to deal with such conflicting visual inputs across the eyes, even binocular rivalry is unrealistic! Sometimes it is fun to investigate those seemingly unrealistic functions of our brains since those may also reveal the mystery of our neural system. As we know, despite binocular rivalry is uncommon in daily life, it is frequently used to investigate awareness. And in our work, we use binocular rivalry to measure perceptual ocular dominance.

Finally, the reviewer queried about if the "dichoptic-backward-movie" adaptation paradigm follow efficient coding and information theory. The information theory and efficient coding assume that messages with low expectedness or of rare occurrence would attract more attention and induce larger neural responses than those with high expectedness. In the "dichoptic-backward-movie" adaptation paradigm, the backward movie should be less expected since the actions of the characters in the backward movie appeared illogical. Thus, according to the information theory and efficient coding, it would be expected that more attention was paid to the backward movie and thus the backward movie might dominate the awareness for a longer period during adaptation (Zhang et al., 2012). However, we instructed participants to follow the regular movie during adaptation. The results of blob detection task also showed a better task performance when the targets appeared in the eye presented with the regular movie, which contradicted with the prediction of the information theory and efficient coding. Thus, it seems not very likely that the "dichoptic-backward-movie" adaptation followed efficient coding and information theory.

References

Bai, J., Dong, X., He, S., & Bao, M. (2017). Monocular deprivation of Fourier phase information boosts the deprived eye’s dominance during interocular competition but not interocular phase combination. Neuroscience, 352, 122-130. https://doi.org/10.1016/j.neuroscience.2017.03.053

Benjamini, Y., & Hochberg, Y. (1995). Controlling the false discovery rate: a practical and powerful approach to multiple testing. Journal of the Royal statistical society: series B (Methodological), 57(1), 289-300. https://doi.org/10.1111/j.2517-6161.1995.tb02031.x

Choe, E., & Kim, M.-S. (2022). Eye-specific attentional bias driven by selection history. Psychonomic Bulletin & Review, 29(6), 2155-2166. https://doi.org/10.3758/s13423-022-02121-0

Corbetta, M., & Shulman, G. L. (2002). Control of goal-directed and stimulus-driven attention in the brain. Nature reviews neuroscience, 3(3), 201-215. https://doi.org/10.1038/nrn755

Dong, X., Gao, Y., Lv, L., & Bao, M. (2016). Habituation of visual adaptation. Sci Rep, 6, 19152. https://doi.org/10.1038/srep19152

Duecker, F., Formisano, E., & Sack, A. T. (2013). Hemispheric differences in the voluntary control of spatial attention: direct evidence for a right-hemispheric dominance within frontal cortex. Journal of Cognitive Neuroscience, 25(8), 1332-1342. https://doi.org/10.1162/jocn_a_00402

Esterman, M., Liu, G., Okabe, H., Reagan, A., Thai, M., & DeGutis, J. (2015). Frontal eye field involvement in sustaining visual attention: evidence from transcranial magnetic stimulation. Neuroimage, 111, 542-548. https://doi.org/10.1016/j.neuroimage.2015.01.044

Gallotto, S., Schuhmann, T., Duecker, F., Middag-van Spanje, M., de Graaf, T. A., & Sack, A. T. (2022). Concurrent frontal and parietal network TMS for modulating attention. iScience, 25(3), 103962. https://doi.org/10.1016/j.isci.2022.103962

Lega, C., Ferrante, O., Marini, F., Santandrea, E., Cattaneo, L., & Chelazzi, L. (2019). Probing the neural mechanisms for distractor filtering and their history-contingent modulation by means of TMS. Journal of Neuroscience, 39(38), 7591-7603. https://doi.org/10.1523/JNEUROSCI.2740-18.2019

Lunghi, C., Burr, D. C., & Morrone, C. (2011). Brief periods of monocular deprivation disrupt ocular balance in human adult visual cortex. Curr Biol, 21(14), R538-539. https://doi.org/10.1016/j.cub.2011.06.004

Lyu, L., He, S., Jiang, Y., Engel, S. A., & Bao, M. (2020). Natural-scene-based Steady-state Visual Evoked Potentials Reveal Effects of Short-term Monocular Deprivation. Neuroscience, 435, 10-21. https://doi.org/10.1016/j.neuroscience.2020.03.039

Mayrhofer, H. C., Duecker, F., van de Ven, V., Jacobs, H. I., & Sack, A. T. (2019). Hemifield-specific correlations between cue-related blood oxygen level dependent activity in bilateral nodes of the dorsal attention network and attentional benefits in a spatial orienting paradigm. Journal of Cognitive Neuroscience, 31(5), 625-638. https://doi.org/10.1162/jocn_a_01338

Rezec, A., Krekelberg, B., & Dobkins, K. R. (2004). Attention enhances adaptability: evidence from motion adaptation experiments. Vision Res, 44(26), 3035-3044. https://doi.org/10.1016/j.visres.2004.07.020

Sack, A. T. (2010). Using non-invasive brain interference as a tool for mimicking spatial neglect in healthy volunteers. Restorative neurology and neuroscience, 28(4), 485-497. https://doi.org/10.3233/RNN-2010-0568

Said, C. P., & Heeger, D. J. (2013). A model of binocular rivalry and cross-orientation suppression. PLoS computational biology, 9(3), e1002991. https://doi.org/10.1371/journal.pcbi.1002991

Song, F., Lyu, L., Zhao, J., & Bao, M. (2023). The role of eye-specific attention in ocular dominance plasticity. Cerebral Cortex, 33(4), 983-996. https://doi.org/10.1093/cercor/bhac116

van den Bergh, D., Wagenmakers, E.-J., & Aust, F. (2023). Bayesian Repeated-Measures Analysis of Variance: An Updated Methodology Implemented in JASP. Advances in Methods and Practices in Psychological Science, 6(2), 25152459231168024. https://doi.org/10.1177/25152459231168024

van Doorn, J., van den Bergh, D., Böhm, U., Dablander, F., Derks, K., Draws, T., Etz, A., Evans, N. J., Gronau, Q. F., Haaf, J. M., Hinne, M., Kucharský, Š., Ly, A., Marsman, M., Matzke, D., Gupta, A., Sarafoglou, A., Stefan, A., Voelkel, J. G., & Wagenmakers, E. J. (2021). The JASP guidelines for conducting and reporting a Bayesian analysis. Psychonomic Bulletin & Review, 28(3), 813–826. https://doi.org/10.3758/s13423-020-01798-5

Wagenmakers, E. J., Love, J., Marsman, M., Jamil, T., Ly, A., Verhagen, J., Selker, R., Gronau, Q. F., Dropmann, D., Boutin, B., Meerhoff, F., Knight, P., Raj, A., van Kesteren, E. J., van Doorn, J., Šmíra, M., Epskamp, S., Etz, A., Matzke, D., de Jong, T., van den Bergh, D., Sarafoglou, A., Steingroever, H., Derks, K., Rouder, J. N., & Morey, R. D. (2018). Bayesian inference for psychology. Part II: Example applications with JASP. Psychonomic Bulletin & Review, 25(1), 58–76. https://doi.org/10.3758/s13423-017-1323-7

Watanabe, M., Cheng, K., Murayama, Y., Ueno, K., Asamizuya, T., Tanaka, K., & Logothetis, N. (2011). Attention but not awareness modulates the BOLD signal in the human V1 during binocular suppression. Science, 334(6057), 829-831. https://doi.org/10.1126/science.1203161

Wong, S. P., Baldwin, A. S., Hess, R. F., & Mullen, K. T. (2021). Shifting eye balance using monocularly directed attention in normal vision. J Vis, 21(5), 4. https://doi.org/10.1167/jov.21.5.4

Yuval-Greenberg, S., & Heeger, D. J. (2013). Continuous flash suppression modulates cortical activity in early visual cortex. J Neurosci, 33(23), 9635-9643. https://doi.org/10.1523/jneurosci.4612-12.2013

Zhang, P., Jiang, Y., & He, S. (2012). Voluntary attention modulates processing of eye-specific visual information. Psychol Sci, 23(3), 254-260. https://doi.org/10.1177/0956797611424289

Zhou, J., Reynaud, A., & Hess, R. F. (2014). Real-time modulation of perceptual eye dominance in humans. Proc Biol Sci, 281(1795). https://doi.org/10.1098/rspb.2014.1717

Version 2

Abstract

Abstract

Previous research has found that prolonged eye-based attention can bias ocular dominance. If one eye long-termly views a regular movie meanwhile the opposite eye views a backward movie of the same episode, perceptual ocular dominance will shift towards the eye previously viewing the backward movie. Yet it remains unclear whether the role of eye-based attention in this phenomenon is causal or not. To address this issue, the present study relied on both the functional magnetic resonance imaging (fMRI) and transcranial magnetic stimulation (TMS) techniques. We found robust activation of the frontal eye field (FEF) and intraparietal sulcus (IPS) when participants were watching the dichoptic movie while focusing their attention on the regular movie. Interestingly, we found a robust effect of attention-induced ocular dominance shift when the cortical function of vertex or IPS was transiently inhibited by continuous theta burst stimulation (cTBS), yet the effect was significantly attenuated to a negligible extent when cTBS was delivered to FEF. A control experiment verified that the attenuation of ocular dominance shift after inhibitory stimulation of FEF was not due to any impact of the cTBS on the binocular rivalry measurement of ocular dominance. These findings suggest that the fronto-parietal attentional network is involved in controlling eye-based attention in the “dichoptic-backward-movie” adaptation paradigm, and in this network, FEF plays a crucial causal role in generating the attention-induced ocular dominance shift.

Introduction

Selective attention allows for the selection of pertinent information from a vast array of input for further cognitive processing (Treisman, 1969; Wolfe et al., 1989). It has been well known that attentional selection can be location-specific (Posner, 1980), feature-specific (Corbetta et al., 1990), or object-specific (Egly et al., 1994). Notably, selective attention can also be based on the eye of origin for visual input, named eye-based attention (Neisser & Becklen, 1975). For instance, presenting top-down attentional cues to one eye can intensify the competition strength of input signals in the attended eye during binocular rivalry (Choe & Kim, 2022; Zhang et al., 2012) and shift the eye balance towards the attended eye (Wong et al., 2021).

It is worth noting that eye-based attention not only exerts real-time effects on information processing (Choe & Kim, 2022; Wong et al., 2021; Zhang et al., 2012), but also exhibits a counterintuitive perceptual aftereffect that prolonged attention to a monocular pathway (or eye-based attention) can result in a shift of ocular dominance towards the unattended eye (Song et al., 2023; Wang et al., 2021). In Song et al. (2023)’s “dichoptic-backward-movie” adaptation paradigm (see Figure 1B), participants were presented with regular movie images in one eye (i.e., attended eye) while the other eye (i.e., unattended eye) received the backward movie images of the same episode. They were also instructed to try their best to follow the logic of the regular movie and ignore the superimposed backward movie. Therefore, the goal-directed eye-based attention was predominantly focused on the attended eye. Song et al. (2023) found that the predominance of the unattended eye in binocular rivalry increased after one hour of adaptation to the “dichoptic-backward-movie”, indicating a shift of perceptual ocular dominance towards the unattended eye. Since the overall energy of visual input from the two eyes was balanced throughout the adaptation period, the change of ocular dominance after adaptation is thought to result from unbalanced eye-based attention rather than unbalanced input energy as in typical short-term monocular deprivation (Bai et al., 2017; Lunghi et al., 2011; Zhou et al., 2014). In short-term monocular deprivation, input signal from one eye is blocked. Accordingly, attention is biased towards the non-deprived eye. However, it is difficult to tease apart the potential contribution of unbalanced eye-based attention from the consequence of the unbalanced input energy, as the deprived eye is also the unattended eye. Therefore, the advantage of the “dichoptic-backward-movie” adaptation paradigm is to balance the input energy across the eyes but leave attention unbalanced across the eyes. Furthermore, the steady-state visually evoked potentials (SSVEPs) elicited by the adapting stimuli in each eye were observed in both the occipital and frontal sites (Song et al., 2023), with a significant positive correlation between the frontal neural-activity index (a normalized SSVEP response related to selective attention) and the shift of perceptual ocular dominance. Their findings imply the contribution of frontal attentional system in producing such an adaptation effect.

Schematic diagram of the (A) binocular rivalry task, (B) “dichoptic-backward-movie” paradigm and (C) blob target stimulus (see the gray region around the mouth in this example). This figure presents identifiable images of human faces solely for the purpose of demonstration, which were captured from the authors (F. Song, J. Wang, and J. Zhao) of this article. The movie images used in the experiment are not displayed in this figure due to potential copyright issues.

However, this account is based on the correlation result rather than causal evidence. Besides, the frontal activation observed by Song et al. (2023) is primarily located in the frontal pole, which is not considered a key area directly responsible for attentional control (Iidaka, 2017; Jääskeläinen et al., 2016; Leminen et al., 2020; Xie et al., 2018). As SSVEP is limited to detecting brain activities that oscillate at the flickering frequency (and its harmonics) of the visual stimulus, brain activities that do not precisely align with the flickering frequency may be missed (Norcia et al., 2015). Consequently, the activities of typical brain regions associated with the attentional network, such as the parietal and dorsolateral prefrontal cortex (Vossel et al., 2014), may not be captured by the SSVEP measures.

Therefore, the present study combines the use of fMRI and TMS to further investigate the neural mechanisms underlying the perceptual aftereffects led by “dichoptic-backward-movie” adaptation, seeking direct causal evidence for the role of eye-based attention in modulating ocular dominance. In Experiment 1, we used fMRI to identify any brain regions associated with eye-based attention when participants were watching the dichoptic movie. Subsequently, in Experiment 2, we examined whether transiently suppressing the cortical function of the identified brain regions using cTBS (Huang et al., 2005) would impair the formation of ocular dominance shift induced by “dichoptic-backward-movie” adaptation. Participants watched the dichoptic movie for 50 minutes after the delivery of cTBS, with their ocular dominance measured before and after adaptation by a binocular rivalry task. The fundamental logic of Experiment 2 can be summarized as follows. Firstly, the brain regions identified in Experiment 1 are related to eye-based attentional control, but do not necessarily contribute to the formation of ocular dominance shift. Secondly, if any of the identified brain regions plays a causal role in the attention-induced ocular dominance plasticity, transiently suppressing its cortical function will sharply weaken the magnitude of ocular dominance shift as compared to delivering the cTBS over vertex, a common control site in TMS studies. Thirdly, for a brain region with a causal role in generating the ocular dominance shift, whether suppression of its cortical function can completely eliminate the effect of ocular dominance shift or not may depend on the depth or efficiency of the suppression led by cTBS. In other words, even if the effect of ocular dominance shift does not vanish after the brain region is suppressed, as long as the effect becomes weakened, the brain region is still thought to play a causal role in the attention-induced ocular dominance plasticity. Experiments 3 and 4 further confirmed that the delivery of cTBS affected the formation of attention induced ocular dominance plasticity during adaptation rather than either the performance of binocular rivalry task or the visibility of monocular stimuli.

Results

Experiment 1: The function role of fronto-parietal areas in controlling eye-based attention

We first tried to identify any brain regions associated with eye-based attention using fMRI in Experiment 1. During the experimental runs of fMRI scanning, participants viewed movie images either dichoptically (Figure 2A) or binocularly (Figure 2B) through the paper red-blue stereoscopic glasses. In the dichoptic condition, regular movie images were presented to one eye while the other eye received the backward movie images of the same episode. In the binocular condition, the same regular movie images were presented to both eyes. Participants were instructed to attentively watch the regular movie and comprehend the plot. Besides, we compared the response difference between dichoptic and binocular conditions in the experimental runs with that in the control runs, where participants performed a central rapid serial visual presentation (RSVP) task while watching the same movie stimuli as those of the experimental runs (Figure 2C), to eliminate potential interference arising from different visual inputs between the dichoptic and binocular conditions.

The schematic of stimuli in (A) dichoptic condition or (B) the binocular condition in the experimental runs. (C) The schematic of RSVP task in the control runs. For demonstration purpose, the letters and fixation points are enlarged. (D) Schematic illustration of the generation process for red-blue movie images employed in the dichoptic condition.

Behavior result

After each experimental run, participants were required to answer a question about the movie plot. The results showed a high response accuracy to the questions (M = 91.67%, SD = 14.91%), indicating that the participants had paid close attention to the regular movie and clearly understood the plot.

In the control runs, participants also exhibited a reasonable performance in the RSVP task (d’ = 3.33, SD = 0.83). A 2 (experiment condition: dichoptic, binocular) × 3 (run: run1, run2, run3) repeated measurements analysis of variance (ANOVA) was conducted on the d-prime score, which indicated no significant main effect or interaction (ps > .09). The generally favorable RSVP performance across both experimental conditions suggested that participants faithfully focused their attention on the central letter stream without being much affected by the movie stimuli in the control runs. Otherwise, the RSVP performance would be found better in the binocular condition than in the dichoptic condition, because a disobedient participant who liked to complete the RSVP task and watch movie simultaneously had to pay more attention to the regular movie images (thus less attention to the letter stream) in the dichoptic condition than in the binocular condition given the annoying bistable nature of dichoptic stimulus presentation.

fMRI Result

It was only in the dichoptic condition of experimental runs that participants had to selectively pay more attention to one eye (i.e., eye-based attention). Therefore, we speculate that if certain brain regions exhibit greater activities in the dichoptic condition as compared to the binocular condition in the experimental runs but not in the control runs, the activation of these brain regions could be attributable to eye-based attention. To seek these brain regions, we used the AFNI program “3dttest++” to access the difference of ‘dichoptic-binocular’ contrast between the experimental and control runs. The AFNI program “ClustSim” was then applied for multiple comparison correction, yielding a minimum significant cluster size of 21 voxels (voxel wise p = .001; cluster threshold α = 0.05). We found 4 clusters showing stronger responses to the dichoptic movies than to the binocular movies especially in the experimental runs (Figure 3). They located in the right FEF, bilateral IPS, and the right superior parietal lobule (SPL). In the control runs, the responses of these areas to the dichoptic and binocular movies showed no discernible difference. As previous research has demonstrated that these brain areas are part of the dorsal attentional network (Vossel et al., 2014), we therefore speculate that the identified clusters would be responsible for eye-based attention.

Illustration of the clusters with stronger ‘dichoptic-binocular’ contrast in the experimental runs than in the control runs. They were located in the right frontal eye field (FEF), bilateral intraparietal sulcus (IPS), and the right superior parietal lobule (SPL). The green circles indicate the corresponding cluster. The MNI coordinates represent the locations of peak voxels of each cluster. The color bar denotes the difference of ‘dichoptic-binocular’ contrast (β values) between the experimental and control runs.

Experiment 2: Suppressing FEF with cTBS attenuated the ocular dominance shift induced by eye-based attention

The results of Experiment 1 suggest that areas within the fronto-parietal network are responsible for eye-based attention. To further ascertain the causal role of eye-based attention in reshaping the ocular dominance, we then measured the effect of “dichoptic-backward-movie” adaptation (Song et al., 2023) on ocular dominance with the function of any identified fronto-parietal area inhibited by cTBS. If suppressing any of the fronto-parietal regions results in a reduction or disappearance of ocular dominance shift after adaptation, we can infer that eye-based attention plays a causal role in the so-called attention-induced ocular dominance plasticity (Song et al., 2023). Participants completed a binocular rivalry task for the measurement of their ocular dominance before (pre-test) and after adaptation (post-test) (Figure 4A). cTBS was delivered over one of the areas identified in Experiment 1 (i.e., right FEF, right IPS) or a control region (vertex, (Wang et al., 2020)) after the pre-test (Figure 4B). Then, participants adapted to the “dichoptic-backward-movie” in which regular movie images were presented to the dominant eye to maximize the effect of eye dominance shift (Song et al., 2023). Meanwhile, they were asked to detect some infrequent blob targets presented on the movie images in one eye.

Illustration of (A) the process in Experiment 2 and (B) the stimulation sites of cTBS.

Binocular rivalry test

To quantify perceptual eye dominance, we calculated the ocular dominance index (ODI) based on binocular rivalry dynamics. An increase in the value of ODI signifies a shift of ocular dominance towards the unattended eye. A 2 (test phase: pre-test, post-test) × 3 (stimulation site: Vertex, FEF, IPS) repeated measures ANOVA on the ODI (Figure 5A) showed a significant main effect of test phase (F (1,15) = 5.12, p = .039, η2 = .26) but not the main effect of stimulation site (F (2,30) = 0.24, p = .79, η2 = .02)), suggesting that the ODI in the post-test (M = 0.49, SE = 0.02) was greater than that in the pre-test (M = 0.46, SE = 0.01, 95% CI = [0.002 0.053]).

The results of (A) the ocular dominance index (ODI), (B) the ocular dominance shift effects (ODSE) in Experiment 2, (C) the ODI and (D) the ODSE in Experiment 3. The bars show the grand average data for each condition. The individual data are plotted with gray lines or dots. The dashed gray line represents the absolute balance point for the two eyes (ODI = 0.5). Error bars indicate standard errors of means. * p < .05; ** p < .01; n.s. p > .05.

More importantly, the interaction between the test phase and the stimulation site was significant (F (2,30) = 3.67, p = .038, η2 = .20). Post-hoc test revealed that after adaptation a significant bias of ocular dominance in favor of the unattended eye could be observed if cTBS was delivered to vertex (ODIpre = 0.451±0.011; ODIpost = 0.494 ±0.015; t (15) = 3.76, p = .002, d = 0.94, 95% CI = [0.019 0.067], pFDR = .006) and IPS (ODIpre = 0.461±0.008; ODIpost = 0.486±0.014; t (15) = 2.39, p = .030, d = 0.60, 95% CI = [0.003 0.049], pFDR = .045), but not FEF (ODIpre = 0.463±0.008, ODIpost = 0.477±0.019; t (15) = 0.74, p = .47, d = 0.19, 95% CI = [-0.025 0.051]), indicating a sharp reduction of aftereffect for stimulation at the FEF only.

We further compared the magnitude of ocular dominance shift across different stimulation sites. For this goal, we computed the ocular dominance shift effect (ODSE) by subtracting the ODI in the pre-test from that in the post-test, then compared the ODSE between the stimulation sites using paired sample t-tests. As shown in Figure 5B, the ODSE after stimulating the vertex (M = 0.043, SE = 0.011) was greater than that after stimulating the FEF (M = 0.013, SE = 0.018; t (15) = 2.76, p = .015, d = 0.69, 95% CI = [0.007 0.053], pFDR = .044). No significant difference was found between other pairs (vertex vs. IPS, M = 0.026, SE = 0.011; t (15) = 1.70, p = .11, d = 0.43, 95% CI = [-0.004 0.039], pFDR = .17; FEF vs. IPS, t (15) = -1.05, p = .31, d = 0.26, 95% CI = [-0.038 0.013]).

Blob detection test

For each experimental condition, the target detection rate was calculated by dividing the summed number of detected blob targets by the total number of blob targets. A 2 (eye: attended eye, unattended eye) × 3 (stimulation site: Vertex, FEF, IPS) repeated measures ANOVA on the detection performance showed a significant main effect of eye (F (1,15) = 112.65, p < .001, η2 = .88) but no significant main effect of stimulation site (F (2,30) = 1.79, p = .18, η2 = .11) or the interaction (F (2,30) = 0.26, p = .78, η2 = .02), suggesting that the overall detection performance for the attended eye (M = 0.89, SE = 0.04) was better than that for the unattended eye (M = 0.23, SE = 0.06, 95% CI = [0.53 0.79]). These results confirm that the interocular difference of detection rate unaffected by the cTBS.

Experiment 3: cTBS affected ocular dominance plasticity rather than the task performance reflecting ocular dominance

The results of Experiment 2 support the notion that eye-based attention was the cause for attention-induced ocular dominance plasticity. However, an alternative account is that the significant two-way interaction between test phase and stimulation site did not stem from any persistent malfunction of FEF in modulating ocular dominance, but rather it was due to some abnormality of binocular rivalry measures in the post-test that occurred after stimulation at the FEF only (and not at the other two brain sites). For instance, stimulation at the FEF might simply reduce the ODI measured in the binocular rivalry post-test.

Therefore, we conducted Experiment 3 to examine how suppression of the three target sites would impact binocular rivalry performance, in case that any unknown confounding factors, which were unrelated to adaptation but related to binocular rivalry measures, contributed to the results. The task procedure was similar to that in Experiment 2 except that participants watched identical regular movie images with both eyes and were no longer required to perform the blob detection task. Since there was no dichoptic-backward-movie adaptation in this experiment, we expected no shift of ocular dominance at least for stimulating the vertex.

A 2 (test phase: pre-test, post-test) × 3 (stimulation site: Vertex, FEF, IPS) repeated measures ANOVA on the ODI showed neither a significant main effect nor the interaction (Figure 5C, ps > 0.43). Moreover, in order to further examine these null effects, we conducted the Bayesian repeated measures ANOVA using JASP with default priors and computed inclusion Bayes factors (BFincl) which suggest the evidence for the inclusion of a particular effect calculated across matched models. Specifically, the Bayesian ANOVA provided moderate evidence (BFincl = 0.28) to support the null hypothesis of no main effect of stimulation site and weak evidence (BFincl = 0.39) in support of the null hypothesis of no main effect of test phase. Additionally, the Bayesian ANOVA yielded moderate evidence (BFincl = 0.23) supporting the null hypothesis of no interaction effect between the test phase and the stimulation site. These results suggested that the suppression of the fronto-parietal cortex had no impact on performance in binocular rivalry task.

Experiment 4: Sound elimination did not impair the blob detection performance following the suppression of FEF

The suppression of FEF via cTBS in Experiment 2 did not impair the performance of blob detection for the attended eye. This result seems confusing, as the interocular difference of blob detection rate is presumed to be a coarse estimation of the real-time effect of eye-based attention. One possible explanation is that with the help of synchronized audio in Experiment 2, residual attentional control following the cTBS might still be competent to sustain attention to the regular movie, allowing a sound performance for this relatively easy task (detection targets appeared only once every 5 min in each eye). To examine this explanation, we conducted Experiment 4, wherein the task procedure was similar to that in Experiment 2 except that sound was no longer presented during the dichoptic-backward-movie adaptation and cTBS was only applied to FEF and vertex. If the synchronized audio (with the regular movie) really contributes to the performance of blob detection for the attended eye after the FEF was suppressed, eliminating the sound may lead to close performance for the attended and unattended eyes.

Binocular rivalry test

A 2 (test phase: pre-test, post-test) × 2 (stimulation site: Vertex, FEF) repeated measures ANOVA on the ODI (Figure 6) showed a significant main effect of test phase (F (1,13) = 22.94, p < .001, η2 = .64) but not the main effect of stimulation site (F (1,13) = 2.67, p = .13, η2 = .17)), suggesting that the ODI in the post-test (M = 0.48, SE = 0.01) was greater than that in the pre-test (M = 0.46, SE = 0.01, 95% CI = [0.01 0.03]). More importantly, the interaction between the test phase and the stimulation site was significant (F (1,13) = 12.76, p = .003, η2 = .50). Post-hoc test revealed that after adaptation a significant bias of ocular dominance in favor of the unattended eye could be observed if cTBS was delivered to vertex (ODIpre = 0.454±0.01; ODIpost = 0.489±0.007; t (13) = 5.41, p < .001, d = 1.45, 95% CI = [0.021 0.049], pFDR < .001), but not FEF (ODIpre = 0.459±0.008, ODIpost = 0.465±0.01; t (13) = 1.24, p = .24, d = 0.33, 95% CI = [-0.005 0.018]).

The results of the ocular dominance index (ODI) in Experiment 4. The bars show the grand average data for each condition. The individual data are plotted with gray lines. The dashed gray line represents the absolute balance point for the two eyes (ODI = 0.5). Error bars indicate standard errors of means. ** p < .01; *** p < .001; n.s. p > .05.

Blob detection test

A 2 (eye: attended eye, unattended eye) × 2 (stimulation site: Vertex, FEF) repeated measures ANOVA on the target detection rate showed a significant main effect of eye (F (1,13) = 101.01, p < .001, η2 = .89) but no significant main effect of stimulation site (F (1,13) = 1.60, p = .23, η2 = .11) or the interaction (F (1,13) = 0.10, p = .76, η2 = .01), suggesting that the overall detection performance for the attended eye (M = 0.79, SE = 0.04) was better than that for the unattended eye (M = 0.23, SE = 0.05, 95% CI = [0.44 0.68]). These results indicate that the interocular difference of detection rate remain unaffected by the cTBS even if no sound was presented.

Discussion

The present study investigated the neural mechanisms underlying the shift of ocular dominance induced by “dichoptic-backward-movie” adaptation. By using fMRI-guided TMS, we provide direct causal evidence for the effect of eye-based attention in modulating ocular dominance, and disclose the crucial role of FEF.

To our knowledge, our Experiment 1 for the first time explored the neural mechanisms of top-down eye-based attention. As compared to when participants watched a movie binocularly, we found stronger activation in FEF, IPS and SPL when they watched the same movie presented in one eye while ignoring the backward movie in the other eye. Since focusing on the movie in one eye necessitates more attention allocated to stimuli in that eye, the activation of these fronto-parietal areas likely reflects their functions in controlling eye-based attention. Given that the activated areas belong to the dorsal attentional network (Vossel et al., 2014), we advocate that top-down eye-based attention is also controlled by the dorsal attentional network. Furthermore, the activation for eye-based attention was predominantly localized in the right hemisphere, showing a typical right-hemisphere functional dominance of attentional control (Duecker et al., 2013; Mayrhofer et al., 2019; Sack, 2010).

Studies on the real-time effect of eye-based attention has shown that top-down eye-based attention can promote the attended eye’s dominance (Wong et al., 2021; Zhang et al., 2012). Yet both recent (Song et al., 2023; Wang et al., 2021) and the present work has suggested that prolonged eye-based attention can lead to a counterintuitive perceptual aftereffect, a shift of ocular dominance towards the unattended eye. Interestingly, in our Experiment 2 this aftereffect was significantly attenuated after we temporarily inhibited the cortical function of FEF via cTBS. This finding indicates the crucial role of FEF in the formation of attention-induced ocular dominance shift.

To exclude the possibility that the cTBS on FEF simply affected the binocular rivalry measures, we conducted Experiment 3, the procedure of which resembled that of Experiment 2 except that participants were presented with regular movie images binocularly. This modification eliminated any involvement of eye-based attentional allocation. The results indicated that the cTBS delivered to FEF did not produce any significant changes on the binocular rivalry performance, thus supporting the notion that inhibition of the FEF undermined its modulatory role in ocular dominance plasticity during adaptation, which, in turn, caused a substantial reduction of ocular dominance shift towards the unattended eye.

Although both the FEF and IPS were found to be responsible for top-down eye-based attention in Experiment 1, the delivery of cTBS to IPS did not affect the adaptation-induced ocular dominance shift. Indeed, both areas belong to the dorsal attentional network (Vossel et al., 2014), yet previous researches have shown some functional distinctions between the two areas. For instance, it is found that the FEF plays a pivotal role in filtering out distractors (Lega et al., 2019) and sustaining attention (Esterman et al., 2015), whereas no such effect has been observed in the IPS. Furthermore, stimulation of the FEF (Ruff et al., 2006; Veniero et al., 2021) and IPS (Ruff et al., 2008) can both modulate activities in early visual areas, but produce qualitatively different effects. In this vein, our work reveals a new type of functional distinction between the two areas, showing a unique role of the FEF in attention-induced ocular dominance plasticity.

Then how does FEF regulate the attention-induced ocular dominance shift? Our previous work has found that the aftereffect (for simplicity, hereafter we use aftereffect to denote the attention-induced ocular dominance shift) can be produced only when the adapting stimuli involve adequate interocular competition, and is measurable only when the testing stimuli are not binocularly fused (Song et al., 2023). Given the indispensability of interocular competition, we explained those findings in the framework of the ocular-opponency-neuron model of binocular rivalry (Said & Heeger, 2013). The model suggests that there are some opponency neurons which receive excitatory inputs from monocular neurons for one eye and inhibitory inputs from monocular neurons for the other eye (e.g. AE-UAE opponency neurons receive excitatory inputs from the attended eye (AE) and inhibitory inputs from the unattended eye (UAE)). Then a difference signal is computed so that the opponency neurons fire if the excitatory inputs surpass the inhibitory inputs. Upon activation, the opponency neurons will in turn suppress the monocular neurons which send inhibitory signals to them.

Based on this model, we proposed an ocular-opponency-neuron adaptation account to explain the aftereffect, and pointed out that the attentional system likely modulated the AE-UAE ocular opponency neurons (Song et al., 2023). So why would FEF modulate the AE-UAE opponency neurons? The reason may be two fold. Firstly, understanding the logic during the dichoptic-backward-movie viewing may require filtering out the distracting information (from the unattended eye) and sustaining attention (to the attended eye), which is exactly the role of FEF (Esterman et al., 2015; Lega et al., 2019). Secondly, due to the special characteristics of binocular vision system, filtering the distracting input from the unattended eye may have to rely on the interocular suppression mechanism. According to the ocular-opponency-neuron model, this is achieved by the firing of the AE-UAE opponency neurons that send inhibitory signals to the UAE monocular neurons.

As mentioned previously, the firing of the AE-UAE opponency neurons requires stronger activity for the AE monocular neurons than for the UAE monocular neurons. This is confirmed by the results shown in Figure 8 of Song et al. (2023) that monocular response for the attended eye during the entire adaptation phase was slightly stronger than that for the unattended eye. Accordingly, during adaptation the AE-UAE opponency neurons were able to activate for a longer period thus adapted to a larger extent than the UAE-AE opponency neurons. This would cause the monocular neurons for the unattended eye to receive less inhibition from the AE-UAE opponency neurons in the post-test as compared with the pre-test, leading to a shift of ocular dominance towards the unattended eye. In this vein, the magnitude of this aftereffect should be proportional to the extent of adaptation of the AE-UAE relative to UAE-AE opponency neurons. Attentional enhancement on the AE-UAE opponency neurons is believed to strengthen this aftereffect, as it has been found that attention can enhance adaptation (Dong et al., 2016; Rezec et al., 2004). Inhibition of FEF likely led such attentional modulation to be much less effective. Consequently, the AE-UAE opponency neurons might not have the chance to adapt to a sufficiently larger extent than the UAE-AE opponency neurons, leading to a statistically non-detectable aftereffect in Experiment 2. Therefore, the results of Experiments 2-4 in the present study suggest that within the context of the ocular-opponency-neuron adaptation account, FEF might be the core area to fulfill the attentional modulations on the AE-UAE opponency neurons.

An unresolved issue is why inhibiting the cortical function of FEF did not impair the performance of blob detection task. One potential explanation is that the synchronized audio in Experiment 2 might help increase the length of time that the regular movie dominated awareness. However, the results of Experiment 4 did not support this explanation, in which the performance of blob detection survived from the inhibition of FEF even when silent movies were presented. Although this issue remains to be explored in future work, it does not contradict with our notion of FEF modulating AE-UAE opponency neurons. It should be noted that our notion merely states that FEF is the core area for attentional modulations on activities of AE-UAE opponency neurons. No other role of FEF during the adaptation is assumed here (e.g. boosting monocular responses or increasing conscious level of stimuli in the attended eye). In contrast, according to the most original definition, the blob detection performance serves as an estimation of visibility (or consciousness level) of the stimuli input from each eye, despite the initial goal of adopting this task is to precisely quantify eye-based attention (which might be impractical). Thus, according to our notion, inhibition of FEF does not necessarily lead to deteriorate performance of blob detection. Furthermore, our findings consistently indicated that the visibility of stimuli in the attended eye was markedly superior to that of stimuli in the unattended eye, yet the discrepancy in the SSVEP monocular responses between the two eyes was minimal though it had reached statistical significance (Song et al., 2023). Therefore, blob detection performance in our work may only faithfully reflect the conscious level in each monocular pathway, but it is probably not an appropriate index tightly associated with the attentional modulations on monocular responses in early visual areas. Indeed, previous work has argued that attention but not awareness modulates neural activities in V1 during interocular competition (Watanabe et al., 2011), but see (Yuval-Greenberg & Heeger, 2013). We have noticed and discussed the counterintuitive results of blob detection performance in our previous work (Song et al., 2023). Here, with the new counterintuitive finding that inhibition of FEF did not impair the performance of blob detection, we suspect that blob detection performance in the “dichoptic-backward-movie” adaptation paradigm may not be an ideal index that can be used to accurately quantify eye-based attention.

In summary, the present study for the first time discovers that the fronto-parietal attentional network is involved in controlling eye-based attention in the “dichoptic-backward-movie” adaptation paradigm. Moreover, the present findings provide direct causal evidence for the role of eye-based attention in modulating ocular dominance. This causal role of eye-based attention is distinct from the homeostatic compensation mechanism commonly used to explain the effect of short-term monocular deprivation (Bai et al., 2017; Lunghi et al., 2011; Lunghi et al., 2015; Lyu et al., 2020; Zhou et al., 2015; Zhou et al., 2013).

Materials and Methods

Experiment 1

Participants

A total of twenty participants were recruited to take part in the screening process (see Stimuli and Procedure), 17 successfully passed the screening test and completed an fMRI experiment. Due to excessive head motion during fMRI scanning, the data of one participant were excluded from further analysis. Eventually, data from 16 participants (12 females, aged from 20 to 28 years) were used. The sample size was predetermined based on the previous study (Song et al., 2023). All participants were naive to the experimental hypotheses. They possessed normal or corrected-to-normal visual acuity, and provided informed consent in accordance with the Declaration of Helsinki. All the experiments in this study were approved (H21058, 11/01/2021) by the Institutional Review Board of the Institute of Psychology, Chinese Academy of Sciences.

Apparatus

During the screening process, stimuli were presented on a gamma-corrected 27-inch ASUS VG279QM LCD monitor (Asus, Shanghai, China, 1920 × 1080 pixels’ resolution, 60 Hz refresh rate) with the mean luminance of 30.18 cd/ m2. Participants viewed the stimuli through a paper red-blue stereoscopic glasses at a distance of 100 cm. A chinrest was used to stabilize head position. The experimental stimuli were programmed using MATLAB (The MathWorks, Natick, USA) and Psychtoolbox (Brainard, 1997; Pelli, 1997). A binocular rivalry task was adopted to screen participants for extreme ocular dominance.

During the fMRI scanning, the stimuli were projected onto a flat panel screen (49 × 36.75 cm, 1024 × 768 pixels’ resolution, 60 Hz refresh rate). The screen was gamma-corrected and placed in the front of the MR scanner using an LCD projector. Participants viewed the screen at the distance of 93 cm via the paper red-blue stereoscopic glasses. The experiment was conducted in a dimly lit room.

Stimuli and Procedure

Screen phase

The stimuli were two orthogonal sinusoidal gratings (±45° from vertical, 2° in diameter, 1.5 cpd) overlapped at the center of screen. We carefully adjusted the contrast of two gratings, so that when participants wore the red-blue glasses, they could only see the red grating through the red glass and the blue grating through the blue glass. The contrasts of red and blue gratings were 48% and 80%, respectively. In addition, a high-contrast checkerboard ‘‘frame” (size: 5°× 5°; thick: 0.25°) and a central white fixation point (diameter: 0.04°) were presented to facilitate binocular fusion (Figure 7).

Illustration of binocular rivalry stimuli in the screening process.

The binocular rivalry test consisted of sixteen 1-min trials. Each trial began with a 5-s blank interval, followed by the presentation of the rival gratings for 55 s. Participants were instructed to hold down one of the three keys (Right, Left, or Down Arrow) to report the grating they perceived (clockwise, counterclockwise, or mixed). The orientation of grating presented to each eye was constant within a trial, but was counter-balanced across trials.

fMRI scanning phase

During fMRI scanning, participants viewed movie images either dichoptically or binocularly with the red-blue glasses. In the dichoptic condition, the movie stimuli were two overlapped red and blue movie episodes (Figure 2A). One was played normally and the other was the same movie but played in a backward sequence (Song et al., 2023). The regular movie images were presented to the dominant eye and the backward ones to the non-dominant eye (for the definition of eye dominance, see Experimental Design below). In the binocular condition, the stimuli were regular movie images colored in purple to ensure visible inputs from both eyes (Figure 2B).

To create the colored movie stimuli, the original movie images were firstly transformed into grayscale. In theory, the red (or blue) movie images can then be generated by zeroing the values of Green and Blue (or Red and Green) channels. However, empirically, the contrast of the red movie images was reduced by applying a multiplication factor of 0.6 to the values of Red channel and setting the value of Blue channel to 80 to achieve perfect monocular stimulation in each eye. The red movie images and the blue movie images were then overlapped to generate the red-blue movie images used in the dichoptic condition (Figure 2D). The purple movie images were obtained by first generating the same red and blue regular movie images in a similar methodology as that employed for the aforementioned red-blue movie images, followed by their overlap. To facilitate binocular fusion, a gray fixation point (diameter: 0.33°) positioned at the center of the movie images and a high-contrast checkerboard ‘‘frame” (size: 17.73°× 23.85°; thick: 0.27°) surrounding the movie were presented during stimulation. The frame rate of movies was 30 fps.

A block design was adopted to study the neural mechanisms related to eye-based attention. A movie block lasted for 30 s, alternated with 12-18 s (12, 14, 16, or 18 s) blanks. Each run consisted of 8 movie blocks. The stimuli of each block could be either the dichoptic movie or the binocular movie, with the sequence counter-balanced.

The fMRI experiment included experimental runs and control runs. During an experimental run, participants were instructed to attentively watch the regular movie and comprehend the plot. To facilitate participants’ understanding of the plot, the sound track for the movie was played synchronized with the regular movie images. After the scanning of each run, a question on the movie plot would be presented on the screen. Participants needed to answer the question via button press, allowing us to examine if they had well followed the plot during the scanning. In a control run, participants performed a rapid serial visual presentation (RSVP) task while watching the movie stimuli (Figure 2C). A series of capital letters were presented to both eyes on the center of the screen. Each letter subtended 0.23°× 0.27°, and was presented for 153 ms. Participants’ task was to press the button whenever they found an ‘X’ in the letter stream. Within each block, the letter ‘X’ appeared for a total of three times. Participants received a feedback about their task performance after the completion of each run. To ensure that participants focused their attention on the RSVP task, the sound tracks were not played during the control runs.

Experimental Design

In the screening test, participants performed three sessions of binocular rivalry tests to measure perceptual ocular dominance. In the binocular rivalry tests, at any given time only the grating in one eye reached awareness and remained visible for a while before the grating in the other eye competed for awareness. The dominant eye was defined as the one that showed longer summed phase durations. Because perceptual ocular dominance fluctuated greatly in the first few trials of a day (Suzuki & Grabowecky, 2007), 5 warm-up trials were completed before the formal task session, the data of which were not analyzed. Perceptual ocular dominance in each test was evaluated by computing an ODI with the formula TL/(TL + TR), where TL and TR represented the summed phase durations for perceiving the stimulus in the left eye and right eye, respectively. Extreme ocular dominance can potentially impede accurate measures of attentional effects. For instance, if the dominant eye is excessively strong, participants may effortlessly perceive the regular movie images for an extended duration, which makes voluntary eye-based attention dispensable. Therefore, participants with ODI values greater than 0.67 or less than 0.33 (3 participants in total) were not allowed to participate in the fMRI scanning.

Each participant completed 3 experimental runs and 3 control runs, with the order counter-balanced across participants. Each run consisted of the movie blocks of dichoptic and binocular conditions.

fMRI Data Acquisition and Preprocessing

fMRI data were collected by a Siemens 3T Magnetom Trio scanner with a 20-channel phased array head coil. T1 weighted anatomical images were acquired at the beginning of scanning (repetition time (TR) = 2600 ms, echo time (TE) = 3.02 ms, flip angle (FA) = 8°, field of view (FOV) = 256 mm, slices = 176, voxel size = 1.0 mm × 1.0 mm × 1.0 mm). Functional data were acquired with T2* weighted Echo-planar imaging (EPI) sequence (TR = 2000 ms, TE = 30 ms, FA = 90°, FOV = 200 mm, slices = 32, voxel size = 3.1 mm × 3.1 mm × 3.5 mm).

The first three volumes were discarded for magnetization equilibrium. The fMRI data were preprocessed using Analysis of Functional NeuroImages (AFNI) software (Cox, 1996). The preprocessing steps encompassed slice timing correction, motion correction, aligning the functional image with anatomy, spatial normalization (MNI152 space), spatial smoothing with 4 mm Gaussian kernel (full width at half maximum, FWHM), and scaling each voxel time series to have a mean of 100. Data from one participant was excluded from further analysis due to the head motion of more than 10% of volumes exceeding 0.3 mm (Silvers et al., 2017; Somerville et al., 2013).

fMRI Data Analysis

We performed single-subject general liner model (GLM) using the AFNI program “3dDeconvolve”. To create regressors for each experimental conditions (dichoptic and binocular), the boxcar function representing the duration of each block (30 s) was convolved with the canonical hemodynamic response function (HRF). Volumes were censored if the motion derivatives exceeded 0.3 mm. High-pass filtering and 3 motion parameters were modeled as non-interest regressors to account for drift and residual motion effects. In addition, single-subject contrasts between experiment conditions (dichoptic > binocular) were calculated.

Subsequently, we used AFNI program “3dttest++” to access difference of the ‘dichoptic-binocular’ contrast between the experimental and control runs. To constrict the analysis on positive activations to movie stimuli, a mask was created specifically for the t-test analysis. This mask was generated by taking the union of all positively activated voxels across all participants for each experiment conditions. “ClustSim” was then applied for multiple comparison correction, yielding a minimum cluster size of 21 voxels (voxelwise p = .001; cluster threshold α = 0.05).

Experiment 2

Participants

Twenty-one participants were initially recruited to participate in the experiment. Eighteen of them successfully passed the screening and completed the formal experiment. The specific screening criteria can be found in the Experimental Design and Data Analysis section. Due to the failure of binocular fusion and the lack of concentration during the experiment, data of two participants were excluded from the analysis. Accordingly, the data from 16 participants (12 females, aged from 18 to 28 years) were analyzed, among which 5 participants have participated in Experiment 1.

Apparatus

Visual stimuli of the behavioral task were displayed on a gamma-corrected 27-inch ASUS VG279QM LCD monitor with the mean luminance of 29.64 cd/ m2. The spatial resolution of monitor was 1920 × 1080 pixels, and the refresh rate was 60 Hz. Participants viewed visual stimuli through a mirror stereoscope with their eyes 100 cm away from the monitors. To stabilize their head position and minimize movement, participants were instructed to rest their heads on a chinrest. The behavioral tasks were performed in a dimly lit room.

Stimuli and Procedure

Binocular Rivalry Test

The binocular rivalry stimuli were two achromatic sine-wave grating disks that were oriented orthogonally (±45° from vertical, diameter: 1°, spatial frequency: 3 cpd, 80% Michelson contrast) and presented foveally to each eye, respectively. A central red fixation point (0.07° in diameter) and a high-contrast checkerboard ‘‘frame” (size: 2.5°×2.5°; thickness: 0.25°) were also presented to both eyes to facilitate stable binocular fusion (Figure 1A). The procedure of binocular rivalry test was consistent with that of Experiment 1.

Dichoptic-backward-movie Adaptation

We employed the “dichoptic-backward-movie” paradigm for the adaptation phase as utilized in the previous research (Song et al., 2023). During the 50-min adaptation, participants were presented with two movie images dichoptically (Figure 1B), with one eye viewing the regular movie images while the other eye viewing the corresponding backward movie images. Note that the backward movie was formally identical to the regular one, except for the absence of a coherent plot. Participants were instructed to exert their utmost effort in comprehending the logic of the regular movie, while disregarding the superimposed backward movie. Therefore, top-down selective attention was supposed to be predominantly directed towards the regular movie. The eye that received the regular movie images was referred to as the “attended eye”, while the other eye as the “unattended eye”. The frame rate of movies was 30 fps.

Blob detection task

The blob detection task has been detailedly described in a prior research (Song et al., 2023). During the adaptation phase, participants were required to watch movies while simultaneously detecting a reduction in color saturation within a blob region that was presented exclusively to one eye (Figure 1C). The locations of the blobs were predetermined by the experimenter, so that they always appeared on faces in the movie. Participants were required to promptly press the SPACE key upon detecting any portion of a character’s face turning gray.

To ensure that the primary focus remained on movie watching, the presentation of the blob was infrequent. To determine the timing of the presentation of the blobs, the adaptation period was divided into 20 segments, each lasting for 150 s. A blob would fade in within 0.2 s at a random moment in the middle 50-s of a segment on the movie of one eye. And after 5 s, it faded out within 0.2 s. There would be 10 blobs presented to each eye.

TMS

We utilized the Magstim Rapid2 stimulator equipped with a standard 70 mm figure-of-eight coil to deliver cTBS. The cTBS protocol consisted of 267 bursts, with each burst comprising 3 pulses at 30 Hz and repeating at 6 Hz. In total, there were 801 pulses delivered and the duration of a single cTBS train was 44 s (Cazzoli et al., 2009; Chaves et al., 2012; Dordevic et al., 2022; Nyffeler et al., 2008). To extend the duration of inhibitory effects induced by the cTBS, two cTBS trains with an interval of 15-min were applied over the same stimulation site. This stimulation protocol has been demonstrated to safely extend the duration of cTBS effects to over 2 hours (Goldsworthy et al., 2012; Nyffeler et al., 2006), thereby ensuring the sustained inhibitory effect throughout the adaptation process. The stimulation intensity was set at 80% of the individual’s resting motor threshold (RMT), which was defined as the minimum machine output required to elicit a visible twitch of the left index finger for 50% of pulses.

Given that the dorsal attentional network primarily consists of the FEF and the IPS (Corbetta & Shulman, 2002; Mayrhofer et al., 2019), with a functional right-hemisphere dominance (Duecker et al., 2013; Mayrhofer et al., 2019; Sack, 2010), we selected the right FEF and right IPS from the four clusters identified in Experiment 1 as the target regions for cTBS (Gallotto et al., 2022). The cTBS stimulation sites were localized based on the peak of the group effect observed in Experiment 1 (MNI coordinates: rFEF: 34.5, -1.5, 55.5; rIPS: 31.5, -73.5, 46.5. see Figure 4B). In addition, we applied cTBS on the vertex (MNI coordinates: 0, 0, 75) as a control condition (Wang et al., 2020). Coil position was guided by the Brainsight 2 neuro-navigation software based on individual T1-weighted images.

Experimental design

Before the formal experiment, all participants underwent a 3-7 days of training period (3 tests per day with a 10-minute rest in between) to ensure consistent performance on the binocular rivalry task (Bao et al., 2018). The task procedure of each day was similar to that of the screen test in Experiment 1. One participant withdrew from the study due to difficulties in achieving binocular fusion.

After the practice of binocular rivalry task, participants were also examined for their capacity in allocating eye-based attention (Neisser & Becklen, 1975). They were asked to watch the dichoptic movie while performing the blob detection task (Song et al., 2023). Only those who showed superior blob detection performance for the attended eyes and subjectively reported paying more attention to the regular movie images were eligible to finish the formal experiment. All the participants who participated in this stage successfully passed the screening. The brain structural images of participants who had not previously participated in Experiment 1 were then acquired.

In the formal experiment, participants performed two binocular rivalry tests initially. The first served as a warm-up test that consisted of five trials and was not analyzed (Bai et al., 2017; Bao et al., 2018). The second comprised 16 trials that measured the perceptual ocular dominance prior to adaptation (i.e., pre-test). Following the pre-test, two cTBS trains with an interval of 15 min were applied over one stimulation site. Afterwards, participants completed a 50-min dichoptic-backward-movie adaption. During the adaptation phase, the main task was to comprehend the plot of the movie. Meanwhile, participants were required to detect the infrequent blob targets. The dichoptic movie was played with sound and the audio track always synchronized with the regular movie images to facilitate the allocation of eye-based attention. As shown in the previous work (Song et al., 2023), presenting the regular movie images to the dominant eye can maximize the effect of eye dominance shift. Therefore, here the dominant eye was always selected as the attended eye. After the end of adaptation, a 16-trial binocular rivalry test was immediately conducted as the post-test (Figure 4A). The formal experiment consisted of three sessions, with cTBS delivered to only one of the three stimulation sites in each session. The order of stimulation site was counter-balanced across participants. The sessions were conducted on different days to minimize any potential carryover effect.

Data Analysis

To quantify perceptual eye dominance, we calculated the ODI based on binocular rivalry dynamics. The ODI was derived using the following formula that yielded a value ranging from 0 (indicating complete dominance of the attended eye) to 1 (indicating complete dominance of the unattended eye).

In the above formula, TAE and TUAE denote the cumulative phase durations of perceiving the stimulus in the attended eye and stimulus in the unattended eye, respectively. Participants with ODI values larger than 0.67 or smaller than 0.33 (2 participants in total) were not allowed to participate in the formal experiment.

Statistical analyses were performed using MATLAB. A 3 (stimulation site: Vertex, FEF, IPS) × 2 (test phase: pre-test and post-test) repeated measures ANOVA was used to investigate the effect of cTBS delivery on ocular dominance shift. Moreover, for the blob detection test, the target detection rate of each experimental condition was calculated by dividing the summed number of detected blob targets by the total number of blob targets. Then, a 2 (eye: attended eye, unattended eye) × 3 (stimulation site: Vertex, FEF, IPS) repeated measures ANOVA on the detection performance was performed. Post-hoc tests were conducted using paired t-tests (2-tailed significance level at α = 0.05), and the resulting p-values were corrected for multiple comparisons using the false discovery rate (FDR) method (Benjamini & Hochberg, 1995).

Experiment 3

Participants

Twenty participants were recruited to participate in the experiment, of which two were excluded for extreme interocular imbalance and another two voluntarily withdrew during the screening phase. Accordingly, 16 participants (14 females, aged from 19 to 26 years) completed all the experiments, one of which had participated in Experiment 1.

Stimuli and Procedure

The task procedure was similar to that in Experiment 2 except that participants watched identical regular movie images with both eyes and were no longer required to perform the blob detection task during adaptation.

Data Analysis

In addition to the data analysis in Experiment 2, we complemented the standard inferential approach with the Bayes factor (van den Bergh et al., 2023; van Doorn et al., 2021; Wagenmakers et al., 2018), which allows quantifying the relative evidence that the data provide for the alternative (H1) or null hypothesis (H0). We conducted the Bayesian repeated measures ANOVA using JASP with default priors and computed inclusion Bayes factors (BFincl) which suggest the evidence for the inclusion of a particular effect calculated across matched models. A BF greater than 1 provides support for the alternative hypothesis. Specifically, a BF between 1 and 3 indicates weak evidence, a BF between 3 and 10 indicates moderate evidence, and a BF greater than 10 indicates strong evidence (van Doorn et al., 2021). In contrast, a BF below 1 provides evidence in favor of the null hypothesis.

Experiment 4

Participants

Fifteen participants were recruited to participate in the experiment. One was excluded for extreme interocular imbalance. Eventually, 14 participants (11 females, aged from 18 to 28 years) completed all the experiments.

Stimuli and Procedure

The task procedure was similar to that in Experiment 2 except that the sound was no longer presented during the dichoptic-backward-movie adaptation and cTBS was only applied to FEF and vertex.